---
title: "Day 2 Project: COVID-19 Severity — SOLUTIONS"
subtitle: "Machine Learning with R in the Life Sciences"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

> **INSTRUCTOR COPY** — This notebook contains complete solutions and
> teaching notes. Do NOT distribute to students.

---

# 0. Setup

```{r packages}
library(tidyverse)
library(tidymodels)
library(vip)
```

---

# 1. Load and Explore the Data

```{r load-data}
expr <- read_csv("covid_expression.csv", show_col_types = FALSE)
meta <- read_csv("covid_metadata.csv", show_col_types = FALSE)

dim(expr)   # 126 samples × 5001 columns (sample_id + 5000 genes)
dim(meta)   # 126 samples × 20 columns
```

```{r prepare-data}
ml_data <- expr |>
  inner_join(meta, by = "sample_id")
gene_cols <- expr |> select(-sample_id) |> colnames()

# ---- TASK A: COVID-19 diagnosis (all 126 samples) ----
target_col <- "disease_state"
ml_data <- ml_data |>
  mutate(target = factor(get(target_col),
                         levels = c("non-COVID-19", "COVID-19"))) |>
  select(all_of(gene_cols), target)

# ---- To switch to TASK B, uncomment below and comment out above ----
# target_col <- "icu"
# ml_data <- ml_data |>
#   filter(disease_state == "COVID-19") |>
#   mutate(target = factor(get(target_col), levels = c("no", "yes"))) |>
#   select(all_of(gene_cols), target)

table(ml_data$target)
```

> **TEACHING NOTE:** Task A has 100 COVID vs 26 non-COVID (77:23 ratio).
> This imbalance is moderate — enough to affect metrics but not catastrophic.
> Task B has 50 ICU vs 50 non-ICU — perfectly balanced. Discuss why balanced
> data makes evaluation simpler.

```{r class-balance}
ggplot(ml_data, aes(x = target, fill = target)) +
  geom_bar() +
  labs(title = "Class Distribution", x = "Class", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Set2")
```

```{r pca}
pca_input <- ml_data |> select(-target)

pca_result <- prcomp(pca_input, center = TRUE, scale. = TRUE)
var_explained <- summary(pca_result)$importance[2, 1:2] * 100

pca_df <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  target = ml_data$target
)

ggplot(pca_df, aes(x = PC1, y = PC2, colour = target)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "PCA of Gene Expression",
    x = paste0("PC1 (", round(var_explained[1], 1), "% variance)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "% variance)"),
    colour = "Class"
  ) +
  theme_minimal() +
  scale_colour_brewer(palette = "Set1")
```

> **TEACHING NOTE (Task A):** Expect clear separation in PCA — COVID-19
> patients cluster distinctly, driven largely by adaptive immune activation
> (plasmablast expansion, immunoglobulin production) and to some extent
> innate immune signaling. PC1 captures ~15-25% of variance and largely
> separates the groups. This predicts that ML will achieve high accuracy
> (85-95%).
>
> **TEACHING NOTE (Task B):** Expect weaker separation for ICU vs non-ICU.
> The two groups overlap considerably because severity differences are more
> subtle. This predicts moderate accuracy (70-85%).

---

# 2. Feature Selection

```{r feature-selection}
gene_variances <- ml_data |>
  select(-target) |>
  summarise(across(everything(), var)) |>
  pivot_longer(everything(), names_to = "gene", values_to = "variance") |>
  arrange(desc(variance))

head(gene_variances, 20)
```

> **TEACHING NOTE:** Students should see IFI27 near the top (variance rank ~3).
> This is the most differentially expressed gene in COVID-19 blood — a key
> interferon-stimulated gene. Other expected top hits: DEFA3/4 (neutrophil
> defensins), LTF (lactoferrin), MMP8 (neutrophil collagenase). Discuss
> how high variance often correlates with biological relevance in disease
> comparisons.

```{r filter-rrna}
# Remove ribosomal RNA genes — technical artifacts, not biology
rrna_pattern <- "^RNA\\d+S"
n_rrna <- sum(str_detect(gene_variances$gene, rrna_pattern))

gene_variances_clean <- gene_variances |>
  filter(!str_detect(gene, rrna_pattern))

cat("Ribosomal RNA genes removed:", n_rrna, "\n")
cat("Genes remaining:", nrow(gene_variances_clean), "\n")
```

> **TEACHING NOTE:** This filtering step is critical for biological
> interpretability. Without it, ~6-8 rRNA genes will appear in the top 20
> important features from the random forest, making the results confusing
> to interpret. The model may still perform well (rRNA levels can
> correlate with sample processing batches that happen to align with
> disease groups), but the biological conclusions would be misleading.
> This is one of the most important lessons: **ML doesn't know or care
> whether a feature is biologically meaningful — it just finds whatever
> separates the classes, including technical noise.**

```{r select-top-genes}
n_features <- 500
top_genes <- gene_variances_clean |>
  slice_head(n = n_features) |>
  pull(gene)

ml_reduced <- ml_data |>
  select(target, all_of(top_genes))

cat("Reduced dataset:", nrow(ml_reduced), "samples,", n_features, "genes\n")
```

---

# 3. Set Up the ML Pipeline

```{r recipe}

set.seed(42)
split <- initial_split(ml_reduced, prop = 0.75, strata = target)
train_data <- training(split)
test_data  <- testing(split)

ml_recipe <- recipe(target ~ ., data = train_data) |>
  step_zv(all_of(top_genes)) |>
  step_normalize(all_of(top_genes))

ml_recipe
```

```{r cv-folds}
set.seed(42)
cv_folds <- vfold_cv(
  train_data,
  v = 10,
  strata = target
)
cv_folds
```

> **TEACHING NOTE:** With 126 samples, 10-fold CV gives ~12 samples per
> fold — reasonable. 5-fold would also work (larger test sets, more stable
> per-fold estimates). LOOCV is possible but computationally expensive and
> gives high-variance estimates. Discuss the bias-variance tradeoff of
> different k values.

---

# 4. Train a Random Forest

```{r rf-spec}
rf_spec <- rand_forest(
  trees = 500,
  mtry = round(sqrt(500)),  # ~22 features per split
  min_n = 5
) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")
```

```{r rf-fit}
rf_workflow <- workflow() |>
  add_recipe(ml_recipe) |>
  add_model(rf_spec)

set.seed(42)
rf_results <- fit_resamples(
  rf_workflow,
  resamples = cv_folds,
  metrics = metric_set(accuracy, roc_auc, sensitivity, specificity),
  control = control_resamples(save_pred = TRUE)
)

collect_metrics(rf_results)
```

> **EXPECTED RESULTS (Task A):** Accuracy ~90-95%, AUC ~0.95-0.99.
> The strong interferon signature makes COVID vs non-COVID classification
> straightforward. If students get lower results, check that the target is
> defined correctly.
>
> **EXPECTED RESULTS (Task B):** Accuracy ~70-80%, AUC ~0.75-0.85.
> ICU vs non-ICU is harder because the transcriptomic differences are more
> subtle. This is a great teaching moment about realistic ML performance.

```{r rf-confusion}
rf_preds <- collect_predictions(rf_results)
conf_mat(rf_preds, truth = target, estimate = .pred_class)
```

```{r rf-confusion-plot}
conf_mat(rf_preds, truth = target, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  labs(title = "Random Forest — Confusion Matrix")
```

> **TEACHING NOTE:** For Task A, most errors will be non-COVID misclassified
> as COVID (false positives), since the minority class has fewer training
> examples. Discuss clinical implications: false negatives (missing a COVID
> patient) are more dangerous than false positives (unnecessary isolation).

---

# 5. Train Elastic Net Logistic Regression

```{r lr-spec}
lr_spec <- logistic_reg(
  penalty = 0.01,
  mixture = 0.5  # Elastic net: blend of L1 (lasso) and L2 (ridge)
) |>
  set_engine("glmnet") |>
  set_mode("classification")
```

```{r lr-fit}
lr_workflow <- workflow() |>
  add_recipe(ml_recipe) |>
  add_model(lr_spec)

set.seed(42)
lr_results <- fit_resamples(
  lr_workflow,
  resamples = cv_folds,
  metrics = metric_set(accuracy, roc_auc, sensitivity, specificity),
  control = control_resamples(save_pred = TRUE)
)

collect_metrics(lr_results)
```

```{r lr-confusion}
lr_preds <- collect_predictions(lr_results)
conf_mat(lr_preds, truth = target, estimate = .pred_class)
```


> **TEACHING NOTE:** Elastic net often performs comparably to random forest
> for Task A (strong signal benefits even simple models). For Task B, RF may
> edge ahead because it captures non-linear interactions. The key teaching
> point: when signal is strong, model choice matters less than when signal
> is weak.

```{r compare-models}
rf_metrics <- collect_metrics(rf_results) |> mutate(model = "Random Forest")
lr_metrics <- collect_metrics(lr_results) |> mutate(model = "Elastic Net")

comparison <- bind_rows(rf_metrics, lr_metrics) |>
  select(model, .metric, mean, std_err)

comparison |>
  pivot_wider(names_from = .metric, values_from = c(mean, std_err))
```

```{r compare-plot}
comparison |>
  ggplot(aes(x = .metric, y = mean, fill = model)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = mean - std_err, ymax = mean + std_err),
    position = position_dodge(width = 0.8), width = 0.2
  ) +
  labs(title = "Model Comparison", y = "Metric Value", x = "") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  coord_cartesian(ylim = c(0.5, 1.0))
```

---

# 6. Feature Importance and Biological Interpretation

```{r feature-importance}
final_rf <- rf_workflow |>
  last_fit(split = split)

final_rf |>
  extract_fit_parsnip() |>
  vip(num_features = 20) +
  labs(title = "Top 20 Important Genes (Random Forest)")
```

```{r top-genes}
top_important <- final_rf |>
  extract_fit_parsnip() |>
  vi() |>
  slice_head(n = 20)

print(top_important)
```

> **ACTUAL TOP GENES (Task A — COVID vs non-COVID):**
>
> The top features from the random forest fall into three clear categories.
> Crucially, they are **not** the interferon genes you might expect from
> the literature — and understanding why is a key learning objective.
>
> **1. Cell proliferation genes (top 3):**
>
> - **RRM2** (Ribonucleotide Reductase M2): The #1 most important feature.
>   Essential for DNA synthesis in dividing cells. Reflects actively
>   proliferating lymphocytes — plasmablasts and activated T cells
>   expanding during the adaptive immune response to SARS-CoV-2.
> - **TYMS** (Thymidylate Synthase): Another DNA synthesis enzyme, also
>   marking cell proliferation. Ranks #2.
> - **MYBL2** (MYB Proto-Oncogene Like 2): A cell cycle transcription
>   factor that drives G2/M progression. Ranks #3.
>
> **2. Immunoglobulin/B-cell genes:**
>
> - **JCHAIN** (Joining Chain): Essential for IgA and IgM polymer assembly.
>   A hallmark marker of antibody-secreting plasma cells.
> - **IGLC1, IGLC2, IGLC3** (Immunoglobulin Lambda Light Chains):
>   Antibody light chain genes — their high importance confirms massive
>   B cell/plasmablast expansion in COVID-19 patients.
> - **IGHG1** (Immunoglobulin Heavy Constant Gamma 1): Encodes IgG1, the
>   dominant antibody subclass in anti-SARS-CoV-2 responses.
> - **TNFRSF17/BCMA** (B-Cell Maturation Antigen): Receptor critical for
>   plasma cell survival — directly marks the plasmablast expansion.
>
> **3. Other:**
>
> - **TXNDC5** (Thioredoxin Domain Containing 5): Highly expressed in
>   plasma cells (part of the ER protein-folding machinery needed for
>   massive antibody production).
> - **EPSTI1** (Epithelial Stromal Interaction 1): This one IS interferon-
>   stimulated — so the innate response signal is not entirely absent,
>   just not dominant.
>
> **Key biological narrative:** The ML model reveals that what best
> *discriminates* COVID-19 from other hospitalized patients is the
> **adaptive immune response** — specifically the massive expansion of
> antibody-secreting cells (plasmablasts) and the associated cell
> proliferation. This makes biological sense: non-COVID hospitalized
> patients may also have activated innate immunity (interferon, neutrophils),
> but they lack the specific SARS-CoV-2-driven B cell response.
>
> **TEACHING NOTE — Why not interferon genes?** Students may be confused
> that IFI27, ISG15, MX1 etc. are not at the top, since these are the
> "classic" COVID-19 genes in the literature. The key insight:
>
> 1. **The comparison group matters.** Published studies compare COVID vs
>    *healthy controls*. Our dataset compares COVID vs *other sick patients*.
>    Both groups likely have elevated interferon responses.
> 2. **ML importance ≠ differential expression.** Importance measures
>    discriminative power between YOUR specific classes, not general
>    biological importance.
> 3. **Correlated features split importance.** The interferon genes are
>    highly co-regulated, so their individual importance gets diluted
>    across many correlated features in the random forest.

```{r boxplot-top-gene}
# Visualize the expression of the top gene
top_gene_name <- top_important$Variable[1]

ml_data |>
  ggplot(aes(x = target, y = .data[[top_gene_name]], fill = target)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1.5) +
  labs(
    title = paste0("Expression of ", top_gene_name, " by Class"),
    y = "log2(TPM + 1)",
    x = "Class"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "none")
```

---

# 7. ML Importance vs. Literature — A Critical Comparison

This is one of the most important pedagogical moments in the project.

```{r literature-comparison}
# Genes consistently highlighted in COVID-19 literature
# (based on COVID vs HEALTHY controls, not vs other sick patients)
literature_genes <- c("IFI27", "ISG15", "MX1", "OAS1", "IFITM3",
                      "CCL2", "S100A8", "S100A12", "IL1B")

# Where do these rank in our variance-selected features?
lit_in_features <- gene_variances_clean |>
  mutate(rank = row_number()) |>
  filter(gene %in% literature_genes)

cat("Literature COVID genes — variance ranks:\n")
print(lit_in_features)

# Where do they rank in RF importance?
all_importance <- final_rf |>
  extract_fit_parsnip() |>
  vi() |>
  mutate(importance_rank = row_number())

lit_in_importance <- all_importance |>
  filter(Variable %in% literature_genes)

cat("\nLiterature COVID genes — RF importance ranks:\n")
print(lit_in_importance)
```

```{r importance-comparison-plot}
# Visualize: compare a literature gene vs a top ML gene
comparison_genes <- c("RRM2", "IFI27", "JCHAIN", "ISG15")
comparison_genes_present <- comparison_genes[comparison_genes %in% colnames(ml_data)]

if (length(comparison_genes_present) > 0) {
  ml_data |>
    select(target, all_of(comparison_genes_present)) |>
    pivot_longer(-target, names_to = "gene", values_to = "expression") |>
    mutate(gene = factor(gene, levels = comparison_genes_present)) |>
    ggplot(aes(x = target, y = expression, fill = target)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.2, size = 1) +
    facet_wrap(~gene, scales = "free_y") +
    labs(
      title = "ML Top Genes vs. Literature Genes",
      subtitle = "RRM2 and JCHAIN discriminate better than IFI27 and ISG15 in this comparison",
      y = "log2(TPM + 1)", x = ""
    ) +
    theme_minimal() +
    scale_fill_brewer(palette = "Set2") +
    theme(legend.position = "none")
}
```

> **TEACHING NOTE — Key lesson for students:**
>
> This comparison is the core critical thinking exercise. Walk students
> through the boxplots:
>
> - **RRM2 and JCHAIN** show clear separation between COVID and non-COVID
>   groups — they are strong *discriminators* for this specific comparison.
> - **IFI27 and ISG15** may show overlap between groups because the non-COVID
>   patients (who are also hospitalized and sick) also have activated
>   interferon responses.
>
> The take-home message: **ML feature importance tells you what separates
> your specific classes, not what is "biologically important" in general.**
> This is why domain knowledge + ML results together are more powerful than
> either alone. A biologist who only reads the ML output might miss the
> interferon story entirely; a biologist who only reads the literature might
> build a classifier on the wrong features.
>
> This is also a good moment to discuss **external validation**: if you
> trained this model on a dataset where the control group was healthy donors
> instead of hospitalized patients, the top features would likely shift
> dramatically toward interferon genes.

---

# 8. ROC Curve

```{r roc-curve}
# ROC curve for Random Forest
rf_preds |>
  roc_curve(truth = target, `.pred_COVID-19`, event_level = "second") |>
  autoplot() +
  labs(title = "ROC Curve — Random Forest (Task A)")
```

> **TEACHING NOTE:** For Task A, expect a ROC curve that hugs the upper-left
> corner (AUC > 0.95). For Task B, the curve will be less ideal (AUC ~0.80).
> This is a good moment to review what AUC means intuitively: the probability
> that a randomly chosen positive sample ranks higher than a randomly chosen
> negative sample.

```{r roc-both}
# Compare ROC curves of both models
rf_roc <- rf_preds |>
  roc_curve(truth = target, `.pred_COVID-19`, event_level = "second") |>
  mutate(model = "Random Forest")

lr_roc <- lr_preds |>
  roc_curve(truth = target, `.pred_COVID-19`, event_level = "second") |>
  mutate(model = "Elastic Net")

bind_rows(rf_roc, lr_roc) |>
  ggplot(aes(x = 1 - specificity, y = sensitivity, colour = model)) +
  geom_line(linewidth = 1) +
  geom_abline(lty = 2, colour = "grey50") +
  labs(title = "ROC Curve Comparison", colour = "Model") +
  theme_minimal() +
  scale_colour_brewer(palette = "Set1")
```

---

# 9. Summary and Reflection

> **MODEL ANSWER (Task A):**
>
> Using bulk RNA-seq from peripheral blood leukocytes (GSE157103), we
> classified 126 hospitalized patients as COVID-19 positive (n=100) or
> negative (n=26). Both random forest and elastic net logistic regression
> achieved high accuracy (~90-95%) and AUC (~0.95-0.99), indicating a strong
> transcriptomic signature of COVID-19 infection. Surprisingly, the most
> important features were NOT the interferon-stimulated genes highlighted
> in the literature (IFI27, ISG15), but rather cell proliferation markers
> (RRM2, TYMS, MYBL2) and immunoglobulin genes (JCHAIN, IGLC1/2/3, IGHG1).
> This reflects the adaptive immune response — specifically the massive
> plasmablast expansion unique to SARS-CoV-2 infection — which distinguishes
> COVID from other hospitalized patients more effectively than the innate
> interferon response (which is shared across many infections). This result
> highlights a critical lesson: ML importance identifies what discriminates
> between specific groups, which may differ from what is generally considered
> "important" in the disease biology literature.
>
> **MODEL ANSWER (Task B):**
>
> Among 100 COVID-19 patients, we classified ICU admission (n=50 ICU vs
> n=50 non-ICU) using blood transcriptomics. Both models achieved moderate
> accuracy (~70-80%) and AUC (~0.75-0.85), reflecting the more subtle
> transcriptomic differences between severity levels. Top predictive genes
> may include neutrophil activation markers (S100A8, S100A12, DEFA3) and
> inflammatory mediators (IL1B), consistent with the role of neutrophil-driven
> inflammation in severe COVID-19. The lower performance compared to Task A
> illustrates that disease severity prediction is inherently harder than
> diagnosis — severity exists on a continuum and depends on host factors
> beyond the transcriptome.

---

# Bonus 1: Hyperparameter Tuning

```{r bonus-tuning}
rf_tune_spec <- rand_forest(
  trees = 500,
  mtry = tune(),
  min_n = tune()
) |>
  set_engine("ranger") |>
  set_mode("classification")

rf_tune_wf <- workflow() |>
  add_recipe(ml_recipe) |>
  add_model(rf_tune_spec)

set.seed(42)
rf_tune_results <- tune_grid(
  rf_tune_wf,
  resamples = cv_folds,
  grid = grid_regular(
    mtry(range = c(10, 50)),
    min_n(range = c(2, 15)),
    levels = 5
  ),
  metrics = metric_set(accuracy, roc_auc)
)

autoplot(rf_tune_results) +
  labs(title = "Hyperparameter Tuning Results")
```

```{r best-params}
show_best(rf_tune_results, metric = "roc_auc")
```

> **TEACHING NOTE:** Tuning typically gives marginal improvement for Task A
> (the signal is so strong that default parameters work well). For Task B,
> tuning may provide a 2-5% improvement. This teaches that hyperparameter
> tuning matters more when signal is weak.

---

# Bonus 2: Clinical + Genomic Features

```{r bonus-clinical}
ml_data <- expr |>
  inner_join(meta, by = "sample_id") |>
  mutate(target = factor(get(target_col),
                         levels = c("non-COVID-19", "COVID-19")))

ml_clinical <- ml_data |>
  select(target, all_of(top_genes), age_years, sex, charlson_score) |>
  mutate(
    age = as.numeric(age_years),
    sex = factor(sex),
    charlson_score = as.numeric(charlson_score)
  ) |>
  drop_na(age, sex, charlson_score)

clinical_recipe <- recipe(target ~ ., data = ml_clinical) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

clinical_wf <- workflow() |>
  add_recipe(clinical_recipe) |>
  add_model(rf_spec)

set.seed(42)
cv_clinical <- vfold_cv(ml_clinical, v = 10, strata = target)

clinical_results <- fit_resamples(
  clinical_wf,
  resamples = cv_clinical,
  metrics = metric_set(accuracy, roc_auc)
)

collect_metrics(clinical_results)
```
```{r bonus-clinical-importance}
final_clinical_rf <- clinical_wf |>
  fit(data = ml_clinical)

final_clinical_rf |>
  extract_fit_parsnip() |>
  vip(num_features = 20) +
  labs(title = "Top 20 Important Genes (Random Forest)")
```

> **TEACHING NOTE:** Adding clinical variables to Task B can improve
> performance slightly (APACHE II and SOFA scores are direct indicators of
> ICU-level illness). For Task A, clinical variables add little because the
> gene expression signal is already very strong. This illustrates the concept
> of complementary information — clinical features help when the genomic
> signal is insufficient.

---

*End of solutions notebook.*
