---
title: "Day 2 Project: Predicting COVID-19 Severity from Blood Transcriptomics"
subtitle: "Machine Learning with R in the Life Sciences"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Welcome to the Day 2 Project!

Today you will apply everything you learned yesterday to a **real RNA-Seq
dataset** from COVID-19 patients. The data comes from **GSE157103**: bulk
RNA-seq of peripheral blood leukocytes from 126 hospitalized patients.

You will work through the full ML pipeline: explore the data, preprocess it,
train classifiers, evaluate performance, and interpret the results biologically.

**Choose one of two classification tasks:**

- **Task A:** Predict **COVID-19 diagnosis** (COVID-19 vs. non-COVID-19) —
  a binary classification with clear transcriptomic signal
- **Task B:** Predict **ICU admission** (ICU vs. non-ICU) among COVID-19
  patients only — a harder binary classification with clinical relevance

Both tasks use the same expression data. Task A uses all 126 samples, while
Task B filters to the 100 COVID-19 positive patients.

> **Reference:** Overmyer et al. (2021) "Large-Scale Multi-omic Analysis of
> COVID-19 Severity." *Cell Systems* 12(1):23-40.

---

# 0. Setup

Install and load all packages. **Run this cell first.**

```{r packages}
# Install if needed (uncomment lines you need)
# install.packages(c("tidyverse", "tidymodels", "ranger", "glmnet", "vip"))

library(tidyverse)
library(tidymodels)
library(vip)
```

---

# 1. Load and Explore the Data

## 1.1 Load the files

```{r load-data}
# Load expression data (126 samples × 5000 genes, log2(TPM+1) transformed)
expr <- read_csv("covid_expression.csv", show_col_types = FALSE)

# Load clinical metadata
meta <- read_csv("covid_metadata.csv", show_col_types = FALSE)

# Quick check — how many samples and genes?
dim(expr)
dim(meta)
```

## 1.2 Prepare the ML dataset

We merge expression and metadata, then define the classification target.

```{r prepare-data}
# Merge expression with metadata by sample_id
ml_data <- expr |>
  inner_join(meta, by = "sample_id")
gene_cols <- expr |> select(-sample_id) |> colnames()

# ---- TASK A: COVID-19 diagnosis (all 126 samples) ----
target_col <- "disease_state"
ml_data <- ml_data |>
  mutate(target = factor(___________,
                         levels = _____________)) |>
  select(all_of(gene_cols), target)

# ---- TASK B: ICU admission (COVID-19 patients only, n=100) ----
# Uncomment below and comment out Task A above to switch
# target_col <- "icu"
# ml_data <- ml_data |>
#   filter(disease_state == "COVID-19") |>
#   mutate(target = factor(________-, levels = ____________)) |>
#   select(all_of(gene_cols), target)

# Check class distribution
table(ml_data$target)
```

## 1.3 Explore the data

### 1.3.1 Class balance

**YOUR TURN:** Create a bar plot of the class distribution.

```{r class-balance}
# FILL IN: Create a bar plot showing the number of samples per class
# Hint: ggplot(ml_data, aes(x = ___, fill = ___)) + geom_bar() + ...
______
```

> **Question:** Is the dataset balanced? What challenges might this create?

### 1.3.2 PCA visualization

**YOUR TURN:** Run PCA on the gene expression data and visualize.

```{r pca}
# Step 1: Remove the target columns , so we only have numeric feature columns
pca_input <- _____________

# Step 2: Run PCA
pca_result <- __________


# Step 3: Calculate variance explained
var_explained <- summary(pca_result)$importance[2, 1:2] * 100

# Step 4: FILL IN — Plot PC1 vs PC2 coloured by target
pca_df <- data.frame(_________________)

# FILL IN: Create a scatter plot of PC1 vs PC2
# Hint: ggplot(pca_df, aes(x = ___, y = ___, colour = ___)) + geom_point() + ...
______
```

> **Question:** Do the classes separate in PCA space? What does this suggest
> about how well ML might work on this task?

---

# 2. Feature Selection

We will select the most variable genes to reduce dimensionality.

```{r feature-selection}
# FILL IN: Calculate variance of each gene across samples
gene_variances <- ml_data |>
  select(-target) |>
  summarise(across(everything(), ___)) |>
  pivot_longer(everything(), names_to = "gene", values_to = "variance") |>
  arrange(desc(variance))

# Look at the top 20 most variable genes
head(gene_variances, 20)
```

```{r filter-rrna}
# IMPORTANT: Remove ribosomal RNA genes before feature selection!
# These are technical artifacts from incomplete rRNA depletion, not biology.
# They start with "RNA5S", "RNA28S", "RNA18S", "RNA45S", etc.
rrna_pattern <- "^RNA\\d+S"

gene_variances_clean <- gene_variances |>
  filter(!str_detect(gene, rrna_pattern))

cat("Genes removed as rRNA:", 
    nrow(gene_variances) - nrow(gene_variances_clean), "\n")
cat("Genes remaining:", nrow(gene_variances_clean), "\n")
```

> **Question:** Why do we remove ribosomal RNA genes? What would happen if we
> left them in — would the model still work, and would the results be
> biologically interpretable?

```{r select-top-genes}
# Select the top 500 most variable genes
n_features <- _____
top_genes <- gene_variances_clean |>
  slice_head(n = n_features) |>
  pull(gene)

# Create a reduced dataset
ml_reduced <- ml_data |>
  select(target, all_of(top_genes))

cat("Reduced dataset:", nrow(ml_reduced), "samples,", n_features, "genes\n")
```

> **Question:** Why might using all 5000 genes be problematic? What is the
> curse of dimensionality?

---

# 3. Set Up the ML Pipeline with tidymodels

## 3.1 Define the recipe

```{r recipe}
# FILL IN: Create a tidymodels recipe
# The formula should predict 'target' from all gene columns
# Include normalization steps

set.seed(42)
split <- initial_split(ml_reduced, prop = 0.75, strata = target)
train_data <- training(split)
test_data  <- testing(split)

ml_recipe <- _____________

# Check the recipe
ml_recipe
```

## 3.2 Define cross-validation

```{r cv-folds}
# FILL IN: Create cross-validation folds
# Hint: use vfold_cv() with a reasonable number of folds
# Consider: we have 126 (or 100) samples — what fold count is appropriate?
set.seed(42)
cv_folds <- __________

cv_folds
```

> **Question:** Why do we use stratified cross-validation? What would happen
> without it when classes are imbalanced?

---

# 4. Train a Random Forest

## 4.1 Specify the model

```{r rf-spec}
# FILL IN: Define a random forest specification
rf_spec <- rand_forest(______) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")
```

> **Hint:** A common rule of thumb for `mtry` is sqrt(number of features).
> With 500 features, that is approximately `r round(sqrt(500))`.

## 4.2 Create the workflow and fit

```{r rf-workflow}
rf_workflow <- workflow() |>
  add_recipe(ml_recipe) |>
  add_model(rf_spec)

# Fit with cross-validation
set.seed(42)
rf_results <- fit_resamples(
  rf_workflow,
  resamples = cv_folds,
  metrics = metric_set(accuracy, roc_auc, sensitivity, specificity),
  control = control_resamples(save_pred = TRUE)
)

# FILL IN: Collect and display the metrics
______
```

## 4.3 Examine the confusion matrix

```{r rf-confusion}
# FILL IN: Collect the predictions and create a confusion matrix
rf_preds <- ______________

# Hint: conf_mat(rf_preds, truth = ___, estimate = ___)
______
```

> **Question:** Which type of error is more common — false positives or false
> negatives? In a clinical setting, which error is more dangerous?

---

# 5. Train a Regularized Logistic Regression (Elastic Net)

## 5.1 Specify the model

```{r lr-spec}
# FILL IN: Define a logistic regression specification with elastic net
lr_spec <- logistic_reg(_______) |>
  set_engine("glmnet") |>
  set_mode("classification")
```

> **Think about it:** What is the difference between Lasso (mixture = 1) and
> Ridge (mixture = 0)? Which one performs feature selection?

## 5.2 Create the workflow and fit

```{r lr-workflow}
lr_workflow <- workflow() |>
  add_recipe(ml_recipe) |>
  add_model(lr_spec)

# Fit with cross-validation
set.seed(42)
lr_results <- fit_resamples(
  lr_workflow,
  resamples = cv_folds,
  metrics = metric_set(accuracy, roc_auc, sensitivity, specificity),
  control = control_resamples(save_pred = TRUE)
)

# Collect metrics
collect_metrics(lr_results)
```

```{r lr-confusion}
lr_preds <- collect_predictions(lr_results)
conf_mat(lr_preds, truth = target, estimate = .pred_class)
```

## 5.3 Compare models

```{r compare-models}
# FILL IN: Create a comparison table of RF vs Logistic Regression
rf_metrics <- collect_metrics(rf_results) |> mutate(model = "Random Forest")
lr_metrics <- collect_metrics(lr_results) |> mutate(model = "Elastic Net")

comparison <- bind_rows(rf_metrics, lr_metrics) |>
  select(model, .metric, mean, std_err)

# Display the comparison
comparison |>
  pivot_wider(names_from = .metric, values_from = c(mean, std_err))
```

```{r compare-plot}
comparison |>
  ggplot(aes(x = .metric, y = mean, fill = model)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = mean - std_err, ymax = mean + std_err),
    position = position_dodge(width = 0.8), width = 0.2
  ) +
  labs(title = "Model Comparison", y = "Metric Value", x = "") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  coord_cartesian(ylim = c(0.5, 1.0))
```

> **Question:** Which model performs better? Is the difference meaningful
> given the standard errors?

---

# 6. Feature Importance and Biological Interpretation

## 6.1 Extract important features

```{r feature-importance}
# Fit the final RF model on all data and extract the final metrics
final_rf <- rf_workflow |>
  last_fit(split =_________)

final_rf_metrics <- ______

# Extract and plot variable importance
final_rf |>
  _________________

```

## 6.2 Biological interpretation

```{r top-genes-biology}
# Get the top 20 gene names
top_important <- final_rf |>
  _____________________

print(top_important)
```

```{r boxplot-top-gene}
# Visualize the expression of the top gene
top_gene_name <- top_important$Variable[1]

ml_data |>
  ggplot(aes(x = target, y = .data[[top_gene_name]], fill = target)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1.5) +
  labs(
    title = paste0("Expression of ", top_gene_name, " by Class"),
    y = "log2(TPM + 1)",
    x = "Class"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "none")
```

> **YOUR TURN — Discussion Questions:**
>
> 1. Look at the top genes. Can you group them into functional categories?
>    (Hint: look for immunoglobulin genes, cell cycle genes, and others.)
> 2. Based on published COVID-19 literature, you might *expect* to see
>    interferon-stimulated genes (IFI27, ISG15, MX1) at the top. Are they
>    there? If not, why might ML importance differ from differential expression?
>    (Hint: think about what the comparison group is — these are hospitalized
>    non-COVID patients, not healthy controls.)
> 3. If you see immunoglobulin genes (JCHAIN, IGLC1/2/3, IGHG1), what does
>    this suggest about the immune response that distinguishes COVID-19 from
>    other conditions in hospitalized patients?
> 4. **Critical thinking:** Does ML feature importance tell you which genes
>    are "important in COVID-19", or which genes best *discriminate between
>    your specific classes*? Why is this distinction crucial?
> 5. How could you validate whether these genes are biologically meaningful
>    (beyond ML importance)?

---

# 7. ML Importance vs. Literature — A Critical Comparison

This section is about **critical thinking**: the top ML features are often
*not* the same genes highlighted in published papers. Understanding why is
one of the most important lessons of this project.

```{r literature-comparison}
# These are genes consistently highlighted in COVID-19 literature
# (based on studies comparing COVID patients vs HEALTHY controls)
literature_genes <- c("IFI27", "ISG15", "MX1", "OAS1", "IFITM3",
                      "CCL2", "S100A8", "S100A12", "IL1B")

# Check: are they in the top 20? Top 50? Top 100? Where do they rank?
lit_in_features <- gene_variances_clean |>
  mutate(rank = row_number()) |>
  filter(__________)

cat("Literature COVID genes — variance ranks:\n")
print(lit_in_features)

# Where do they rank in RF importance?
all_importance <- final_rf |>
  extract_fit_parsnip() |>
  vi() |>
  mutate(importance_rank = row_number())

lit_in_importance <- all_importance |>
  filter(Variable %in% literature_genes)

cat("\nLiterature COVID genes — RF importance ranks:\n")
print(lit_in_importance)
```

```{r importance-comparison-plot}
# Visualize: compare a literature gene vs a top ML gene
comparison_genes <- c("RRM2", "IFI27", "JCHAIN", "ISG15")
comparison_genes_present <- comparison_genes[comparison_genes %in% colnames(ml_data)]

if (length(comparison_genes_present) > 0) {
  ml_data |>
    select(target, all_of(comparison_genes_present)) |>
    pivot_longer(-target, names_to = "gene", values_to = "expression") |>
    mutate(gene = factor(gene, levels = comparison_genes_present)) |>
    ggplot(aes(x = target, y = expression, fill = target)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.2, size = 1) +
    facet_wrap(~gene, scales = "free_y") +
    labs(
      title = "ML Top Genes vs. Literature Genes",
      subtitle = "RRM2 and JCHAIN discriminate better than IFI27 and ISG15 in this comparison",
      y = "log2(TPM + 1)", x = ""
    ) +
    theme_minimal() +
    scale_fill_brewer(palette = "Set2") +
    theme(legend.position = "none")
}
```

> **Discussion — Why the discrepancy?**
>
> - The published literature mostly compares COVID patients to **healthy
>   donors**. Our dataset compares COVID to **other hospitalized patients**
>   with respiratory symptoms. The interferon response may be activated in
>   both groups!
> - ML importance measures **discriminative power between your specific
>   classes**, not general "importance in disease." A gene can be critical
>   to COVID biology yet useless for classification if both groups express
>   it similarly.
> - What does this teach us about interpreting ML results in biology?

---

# 8. Model Evaluation — ROC Curve

```{r roc-curve}
# FILL IN: Create an ROC curve for the best model
# Hint: Use the predictions from rf_results or lr_results
_______ |>
  roc_curve(truth = _____, _______, event_level = "second") |>
  autoplot() +
  labs(title = "ROC Curve — ______")


# FILL IN: Plot the ROC curves of both models
rf_roc <- rf_preds |>
  roc_curve(truth = _____, _______, event_level = "second") |>
  mutate(model = "Random Forest")

lr_roc <- lr_preds |>
  roc_curve(truth = _____, ________, event_level = "second") |>
  mutate(model = "Elastic Net")

bind_rows(rf_roc, lr_roc) |>
  ggplot(aes(x = 1 - specificity, y = sensitivity, colour = model)) +
  geom_line(linewidth = 1) +
  geom_abline(lty = 2, colour = "grey50") +
  labs(title = "ROC Curve Comparison", colour = "Model") +
  theme_minimal() +
  scale_colour_brewer(palette = "Set1")
```

---

# 9. Summary and Reflection

Write a brief summary of your results (4–5 sentences):

> YOUR ANSWER:
>
> - What accuracy / AUC did you achieve?
> - Which model worked best?
> - What were the most important genes?
> - What biological conclusions can you draw?
> - What would you do differently with more time?

---

# Bonus Exercises

## Bonus 1: Hyperparameter Tuning

Try tuning the random forest with `tune()` and `tune_grid()`.

```{r bonus-tuning, eval=FALSE}
# FILL IN: Create a tunable RF specification
rf_tune_spec <- rand_forest(______) |>
  set_engine("ranger") |>
  set_mode("classification")

# FILL IN: Create the workflow and tune
rf_tune_wf <- workflow() |>
  add_recipe(ml_recipe) |>
  add_model(rf_tune_spec)

# FILL IN: Define a tuning grid and run
# Hint: tune_grid(rf_tune_wf, resamples = cv_folds, grid = ___)
______
```

```{r best-params}
show_best(rf_tune_results, metric = "roc_auc")
```

## Bonus 2: Include Clinical Variables

Try adding clinical metadata (age, sex, Charlson score) as additional features.

```{r bonus-clinical, eval=FALSE}
ml_data <- expr |>
  inner_join(meta, by = "sample_id") |>
  mutate(target = factor(get(target_col),
                         levels = c("non-COVID-19", "COVID-19")))

# Create a dataset with both gene expression and clinical features
ml_clinical <- ml_data |>
  select(target, all_of(top_genes), age_years, sex, charlson_score) |>
  mutate(
    age = as.numeric(age),
    sex = factor(sex),
    charlson_score = as.numeric(charlson_score)
  ) |>
  drop_na(age, sex, charlson_score)

# FILL IN: Create a recipe that handles both numeric and categorical predictors
# Hint: add step_dummy(all_nominal_predictors()) for the sex variable
clinical_recipe <- ______

clinical_wf <- workflow() |>
  _________

cv_clinical <- vfold_cv(ml_clinical, v = 10, strata = target)

clinical_results <- fit_resamples(
  ___________,
  resamples = ________,
  metrics = metric_set(accuracy, roc_auc)
)

collect_metrics(clinical_results)
```

## Bonus 3: Compare Task A vs Task B

If you have time, try both tasks and compare results. Why might one task be
easier than the other?

---

*End of project notebook — good luck!*
