---
title: "Exercise 1: Meet Your Data"
subtitle: "Machine Learning with R in the Life Sciences"
author: "Konstantin Pelz, Prof. Dr. Markus List"
date: "2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Goal

In this first exercise, you will practice the basic R skills needed for the rest of the course:
loading data, exploring its structure, and creating simple visualizations. We use the
**Palmer Penguins** dataset -- morphological measurements of three penguin species
collected on islands near Antarctica.

**Time:** ~15 minutes

**Corresponds to:** Introduction and Data section of the lecture

---

# 0 -- Setup

Run this cell to load the required packages. If a package is missing, the code will install it
automatically.

```{r}
if (!require(palmerpenguins)) install.packages("palmerpenguins")
if (!require(tidyverse))      install.packages("tidyverse")

library(palmerpenguins)
library(tidyverse)
```

# 1 -- Load and Inspect the Data

The `palmerpenguins` package provides a data frame called `penguins`. Let us get a first
look at it.

```{r}
# This loads the data into your environment
data("penguins")
```

### YOUR TURN 1.1

Use the functions `dim()`, `head()`, and `summary()` to explore the penguins dataset.
Fill in the blanks (replace `___`).

```{r}
# How many rows and columns does the dataset have?
dim(___)

# Show the first 6 rows
head(___)

# Get a statistical summary of each column
summary(___)
```

> **CHECKPOINT:** You should see that the dataset has **344 rows** and **8 columns**. The
> columns include `species`, `island`, `bill_length_mm`, `bill_depth_mm`,
> `flipper_length_mm`, `body_mass_g`, `sex`, and `year`.

### YOUR TURN 1.2

Use `str()` to see the data types of each column. This is important because ML models
treat numeric and categorical (factor) variables differently.

```{r}
str(___)
```

**Discussion question:** Which columns are numeric? Which are categorical? Which column
could be a *target variable* (something you would want to predict)?

---

# 2 -- Handling Missing Data

Real-world datasets almost always contain missing values. Let us check.

```{r}
# Count missing values per column
colSums(is.na(penguins))
```

### YOUR TURN 1.3

Remove rows with missing values using `drop_na()` and store the result in a new variable
called `penguins_clean`. Then check how many rows remain.

```{r}
penguins_clean <- penguins |>
  ___()

# How many rows are left?
dim(penguins_clean)
```

> **CHECKPOINT:** You should have **333 rows** remaining.

---

# 3 -- Basic Data Manipulation with dplyr

The `tidyverse` provides powerful tools for data manipulation. The pipe operator `|>`
chains operations together: the output of one step becomes the input of the next.

### YOUR TURN 1.4

Count how many penguins of each species are in the cleaned dataset. Fill in the blank
with the correct column name.

```{r}
penguins_clean |>
  count(___)
```

> **CHECKPOINT:** Adelie: 146, Chinstrap: 68, Gentoo: 119.

### YOUR TURN 1.5

Use `group_by()` and `summarize()` to calculate the mean body mass for each species.

```{r}
penguins_clean |>
  group_by(___) |>
  summarize(mean_mass = mean(___))
```

---

# 4 -- Visualization with ggplot2

Visualizing data before modeling is essential. ggplot2 works in layers:
`ggplot(data, aes(x, y))` sets up the canvas, then you add geometries like `geom_point()`.

### YOUR TURN 1.6

Create a bar chart showing the number of penguins per species. Fill in the correct
geometry function.

```{r}
ggplot(penguins_clean, aes(x = species, fill = species)) +
  ___() +
  theme_minimal() +
  labs(title = "Number of Penguins per Species",
       x = "Species", y = "Count")
```

### YOUR TURN 1.7

Create a scatter plot of bill length vs. bill depth, colored by species. This shows whether
the species form distinct clusters in this feature space.

```{r}
ggplot(penguins_clean, aes(x = ___, y = ___, color = species)) +
  geom_point(alpha = 0.7, size = 2) +
  theme_minimal() +
  labs(title = "Bill Dimensions by Species",
       x = "Bill Length (mm)", y = "Bill Depth (mm)")
```

### YOUR TURN 1.8 (Stretch Goal)

Create a boxplot of flipper length by species. Boxplots show the distribution (median,
quartiles, outliers) and are useful for comparing groups.

```{r}
ggplot(penguins_clean, aes(x = ___, y = ___, fill = ___)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Flipper Length by Species",
       x = "Species", y = "Flipper Length (mm)")
```

---

# 5 -- Thinking Like a Data Scientist

Before applying machine learning, we need to think about our data in ML terms.

**Discussion questions (no code needed):**

1. If we wanted to predict the **species** of a penguin from its measurements, which columns
   would be the **features** (inputs) and which would be the **target** (output)?

2. Is this a **classification** or a **regression** problem? Why?

3. Looking at your scatter plot from 1.7: do you think a machine learning model could
   separate the three species? Which species pair might be hardest to distinguish?

4. We have 333 samples. Is this a lot or a little for ML? What if we had 33,000 genes
   instead of 4 numeric features?

---

# Summary

In this exercise you practiced:

- Loading and inspecting data with `dim()`, `head()`, `summary()`, `str()`
- Handling missing values with `drop_na()`
- Data manipulation with `count()`, `group_by()`, `summarize()`
- Visualization with `ggplot2`: bar charts, scatter plots, boxplots
- The pipe operator `|>` for chaining operations

These are the same tools you will use for the rest of the course -- just with more complex
data and ML-specific functions on top.

---

# Hints

- 1.3: The function is `drop_na` and it goes inside the pipe.
- 1.5: Group by `species`, summarize with `body_mass_g`.
- 1.6: The geometry for bar charts is `geom_bar`
- 1.7: Use `bill_length_mm` for x and `bill_depth_mm` for y.

---

*Next up: Exercise 2 -- Unsupervised Exploration (PCA and clustering)*
