---
title: "Bonus Exercise: Spot the Leakage"
subtitle: "Machine Learning with R in the Life Sciences"
author: "Konstantin Pelz, Prof. Dr. Markus List"
date: "2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Goal

Data leakage is one of the most common and dangerous mistakes in machine learning.
It leads to **over-optimistic performance estimates** that do not hold up on new data.
In this exercise, you will examine code snippets that contain leakage, identify the
problem, and understand why the corrected version is better.

**Time:** ~10 minutes

**Corresponds to:** Pitfalls section of the lecture

**Reference:** Bernett et al. (2024). "Guiding questions to avoid data leakage in biological
machine learning applications." *Nature Methods*. DOI: 10.1038/s41592-024-02362-y

---

# 0 -- Setup

```{r}
if (!require(tidymodels)) install.packages("tidymodels")
if (!require(tidyverse))  install.packages("tidyverse")

library(tidyverse)
library(tidymodels)

# ---- Same simulated data ----
set.seed(2025)
n_samples <- 200; n_genes <- 50; n_de_genes <- 10
sample_class <- factor(rep(c("tumor", "normal"), each = n_samples / 2))
expression_matrix <- matrix(
  rnorm(n_samples * n_genes, mean = 8, sd = 1.5),
  nrow = n_samples, ncol = n_genes
)
colnames(expression_matrix) <- paste0("gene_", 1:n_genes)
for (i in 1:n_de_genes) {
  expression_matrix[1:(n_samples / 2), i] <-
    expression_matrix[1:(n_samples / 2), i] + rnorm(n_samples / 2, mean = 2, sd = 0.5)
}
gene_data <- as.data.frame(expression_matrix) |> mutate(class = sample_class)
```

---

# Scenario 1: Normalizing Before Splitting

A colleague shows you their preprocessing code. They normalize the data, then split
it into train and test sets.

## Version A (Their code)

```{r}
set.seed(123)

# Step 1: Normalize ALL data
all_data_scaled <- gene_data |>
  mutate(across(starts_with("gene_"), scale))

# Step 2: THEN split
split_A <- initial_split(all_data_scaled, prop = 0.75, strata = class)
train_A <- training(split_A)
test_A  <- testing(split_A)
```

## Version B (Alternative)

```{r}
set.seed(123)

# Step 1: Split FIRST
split_B <- initial_split(gene_data, prop = 0.75, strata = class)
train_B <- training(split_B)
test_B  <- testing(split_B)

# Step 2: Normalize using ONLY training data statistics
# (tidymodels recipes handle this automatically)
recipe_B <- recipe(class ~ ., data = train_B) |>
  step_normalize(all_numeric_predictors())
```

### YOUR TURN B.1

**Which version contains data leakage? Explain why.**

Write your answer here:

> *Your answer:* ___
>
>
>

### YOUR TURN B.2

**Why does it matter?** Think about what `scale()` computes (the mean and standard
deviation). Where does the information from the test set "leak" into the training process?

> *Your answer:* ___
>
>
>

---

# Scenario 2: Feature Selection Before Cross-Validation

Another colleague selects the most informative genes before running cross-validation.

## Version A (Their code)

```{r}
set.seed(123)

# Step 1: Find the top 10 genes by t-test across ALL training data
split_2 <- initial_split(gene_data, prop = 0.75, strata = class)
train_2 <- training(split_2)

# Run t-tests on the full training set
p_values <- sapply(paste0("gene_", 1:50), function(g) {
  t.test(train_2[[g]] ~ train_2$class)$p.value
})
top_genes <- names(sort(p_values))[1:10]
cat("Selected genes:", top_genes, "\n")

# Step 2: Subset to top genes, THEN do cross-validation
train_selected <- train_2 |> select(all_of(top_genes), class)

cv_A <- vfold_cv(train_selected, v = 5, strata = class)

rf_wf <- workflow() |>
  add_recipe(recipe(class ~ ., data = train_selected)) |>
  add_model(rand_forest(trees = 500) |> set_mode("classification") |> set_engine("ranger"))

results_A <- rf_wf |> fit_resamples(resamples = cv_A)
cat("\nResults with pre-selected genes:\n")
results_A |> collect_metrics()
```

## Version B (Alternative)

```{r}
set.seed(123)

split_2B <- initial_split(gene_data, prop = 0.75, strata = class)
train_2B <- training(split_2B)

# Feature selection happens INSIDE the recipe (and thus inside CV)
recipe_2B <- recipe(class ~ ., data = train_2B) |>
  step_normalize(all_numeric_predictors()) |>
  step_corr(all_numeric_predictors(), threshold = 0.9)

cv_B <- vfold_cv(train_2B, v = 5, strata = class)

rf_wf_B <- workflow() |>
  add_recipe(recipe_2B) |>
  add_model(rand_forest(trees = 500) |> set_mode("classification") |> set_engine("ranger"))

results_B <- rf_wf_B |> fit_resamples(resamples = cv_B)
cat("\nResults with recipe-based feature handling:\n")
results_B |> collect_metrics()
```

### YOUR TURN B.3

**Which version contains data leakage? Explain the problem.**

> *Your answer:* ___
>
>
>

### YOUR TURN B.4

**Why is Version A particularly dangerous for publication?**

Think about what happens when the selected genes are used on completely new data
from a different hospital or cohort.

> *Your answer:* ___
>
>
>

---

# Scenario 3: The Sneaky Shortcut

A student is predicting tumor vs. normal from gene expression data. Their model
achieves 99.8% accuracy. Suspiciously good!

```{r}
set.seed(2025)

# Their dataset includes a hidden problem
sneaky_data <- gene_data |>
  mutate(
    batch = ifelse(class == "tumor", "batch_1", "batch_2"),
    measurement_time = ifelse(class == "tumor",
                              rnorm(n(), mean = 9, sd = 0.5),   # morning
                              rnorm(n(), mean = 15, sd = 0.5))  # afternoon
  )

cat("Preview of the data:\n")
sneaky_data |> select(class, batch, measurement_time, gene_1, gene_2) |> head(10)
```

### YOUR TURN B.5

**What is wrong with this dataset? Why would a model achieve near-perfect accuracy,
and why would this NOT work on new data?**

> *Your answer:* ___
>
>
>

### YOUR TURN B.6

**Name two types of information in this dataset that the model could use as shortcuts
instead of learning the actual biological signal.**

> *Your answer:* ___
>
>
>

---

# A Leakage Checklist

When conducting ML in your research, ask yourself these questions (adapted from
Bernett et al. 2024):

1. **Was any preprocessing done on the full dataset before splitting?**
   Normalization, feature selection, imputation, and outlier removal should all
   happen AFTER splitting (or inside the CV loop via recipes).

2. **Could any feature serve as a proxy for the label?**
   Batch IDs, measurement times, hospital codes, or technician IDs that correlate
   with the outcome will give misleadingly high performance.

3. **Are there duplicate or near-duplicate samples?**
   Technical replicates of the same patient in both train and test sets will inflate
   performance. Split by patient ID, not by sample.

4. **Was feature selection done inside or outside the CV loop?**
   If outside, the model has already "peeked" at the test fold information through
   the feature selection step.

5. **Is the test set performance suspiciously close to the CV performance?**
   Some gap is normal. If test performance is much higher than expected,
   investigate potential leakage.

---

# Summary

Data leakage means that information from the test set (or future data) influences the
training process. Common forms include:

- Normalizing or preprocessing before splitting
- Selecting features on the full dataset before CV
- Confounding variables (batch effects, measurement artifacts) that correlate with
  the target
- Duplicate samples across train and test

**Key takeaway:** If your model seems too good to be true, it probably is. Always check
for leakage before trusting your results.

---

*This concludes the hands-on exercises for Day 1. Tomorrow you will apply everything
you learned in a group project on a real transcriptomics dataset!*
