---
title: "Exercise 3: Your First Classifier -- SOLUTIONS"
subtitle: Machine Learning with R in the Life Sciences
output:
  pdf_document:
    toc: false
  html_document:
    toc: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Setup

```{r}
library(tidyverse)
library(tidymodels)

set.seed(2025)
n_samples <- 200; n_genes <- 50; n_de_genes <- 10
sample_class <- factor(rep(c("tumor", "normal"), each = n_samples / 2))
expression_matrix <- matrix(rnorm(n_samples * n_genes, mean = 8, sd = 1.5),
                            nrow = n_samples, ncol = n_genes)
colnames(expression_matrix) <- paste0("gene_", 1:n_genes)
for (i in 1:n_de_genes) {
  expression_matrix[1:(n_samples / 2), i] <-
    expression_matrix[1:(n_samples / 2), i] + rnorm(n_samples / 2, mean = 2, sd = 0.5)
}
gene_data <- as.data.frame(expression_matrix) |> mutate(class = sample_class)
```

### 3.1
```{r}
ggplot(gene_data, aes(x = class, y = gene_1, fill = class)) +
  geom_boxplot() + theme_minimal() + labs(title = "Gene 1 (Differentially Expressed)")

ggplot(gene_data, aes(x = class, y = gene_40, fill = class)) +
  geom_boxplot() + theme_minimal() + labs(title = "Gene 40 (Not Differentially Expressed)")
```

**Q: Can you visually see the difference in gene_1? What about gene_40?**

**gene_1:** Yes — the boxplot should show a clear shift between the tumor and normal
groups. This is expected because gene_1 is one of the 10 truly differentially
expressed (DE) genes in our simulation. The tumor group has a noticeably higher (or
lower, depending on the simulation run) median expression, and the boxes have
limited overlap.

**gene_40:** No meaningful visual difference. The two boxplots should look nearly
identical — similar medians, similar spread, heavy overlap. This is because gene_40
is a noise gene with no built-in association with the class label. Any small difference
you see is just random variation.

This comparison illustrates a fundamental point: **not all features are informative**.
In a real transcriptomics dataset with 20,000 genes, only a small subset will be
relevant to your phenotype of interest. The challenge of ML is to find and use those
informative features while ignoring the noise — which is exactly what feature
selection and well-regularized models do.

It also shows why exploratory visualization is valuable before modeling: a quick look
at a few features can already tell you whether there is signal in your data.

### 3.2
```{r}
set.seed(123)
split <- initial_split(gene_data, prop = 0.75, strata = class)
train_data <- training(split)
test_data  <- testing(split)
cat("Training:", nrow(train_data), "Test:", nrow(test_data), "\n")
table(train_data$class)
```

### 3.3
```{r}
log_reg_spec <- logistic_reg() |>
  set_mode("classification") |>
  set_engine("glm")
```

### 3.4
```{r}
train_recipe <- recipe(class ~ ., data = train_data)
wf <- workflow() |>
  add_recipe(train_recipe) |>
  add_model(log_reg_spec)
fitted_model <- wf |> fit(data = train_data)
```

### 3.5
```{r}
predictions <- fitted_model |>
  predict(new_data = test_data) |>
  bind_cols(test_data |> select(class))
head(predictions)
```

### 3.6
```{r}
conf_mat(predictions, truth = class, estimate = .pred_class)
```

### 3.7
```{r}
predictions |> metrics(truth = class, estimate = .pred_class)
```

### Discussion Questions — Solutions

**Q1: Is accuracy alone sufficient to judge this model? What if the dataset were 95% normal and 5% tumor?**

No. In our simulated dataset we have a balanced 50/50 split (100 tumor,
100 normal), so accuracy is a reasonable metric here. But if the dataset were
**95% normal and 5% tumor**, a trivial model that predicts "normal" for every sample
would achieve 95% accuracy while missing every single tumor case — which is
clinically useless.

In that scenario you must look at:

- **Sensitivity** — what fraction of actual tumors did we detect?
- **Specificity** — what fraction of healthy samples did we correctly identify?
- Metrics designed for imbalance: **balanced accuracy**, **F1 score**, or **AUC**

This is why the confusion matrix is so important: it tells you *where* the errors are,
not just how many.

**Q2: Looking at the confusion matrix: does the model make more false positives or false negatives? Which error is worse in a clinical screening context?**

*(The exact numbers dependon your run, but typically the model makes
both types of errors in roughly similar numbers with this balanced simulated data.)*

In a **clinical screening context**, **false negatives are worse**: a false negative means
a tumor patient is told they are healthy and may not receive timely treatment. A
false positive (healthy person flagged as tumor) leads to unnecessary follow-up
tests, which is stressful and costly but not life-threatening.

General principle in medical screening: we prefer **high sensitivity** (catching as many
true cases as possible) even at the expense of lower specificity (more false alarms).
The acceptable tradeoff depends on the cost and invasiveness of follow-up — for a
cheap blood test you can tolerate many false positives, but for an invasive biopsy
you want fewer.

**Q3: We used ALL 50 genes as features. Some are just noise. Would the model perform better with fewer, carefully selected genes?**

Very likely yes. Only **gene_1 through gene_10** are truly differentially expressed —
the other 40 genes are pure noise. Including noise features forces the model to
"learn" that they are irrelevant, which wastes model capacity and can cause
overfitting, especially with small sample sizes.

**Feature selection** (e.g., `step_nzv()` for near-zero variance filtering,
`step_corr()` for correlation filtering, or model-based importance from a random
forest) could remove the noise genes and likely improve performance.

Key lesson for life science ML: with omics data, feature selection is often more
impactful than choosing a fancier model.

**Important caveat:** Feature selection must happen **inside** the cross-validation loop
to avoid data leakage.