---
title: "Exercise 4: Model Comparison with Cross-Validation"
subtitle: "Machine Learning with R in the Life Sciences"
author: "Konstantin Pelz, Prof. Dr. Markus List"
date: "2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Goal

In Exercise 3, you trained a single model and evaluated it once on the test set. But how
do you know this result is not just lucky? And how do you compare two different models
fairly? In this exercise you will use **cross-validation** to get robust performance
estimates and compare **logistic regression** vs. **random forest** on the same data.

**Time:** ~15 minutes

**Corresponds to:** Model Selection + Cross-Validation sections of the lecture

---

# 0 -- Setup and Data

We reuse the simulated transcriptomics dataset from Exercise 3.

```{r}
if (!require(tidymodels)) install.packages("tidymodels")
if (!require(tidyverse))  install.packages("tidyverse")
if (!require(ranger))     install.packages("ranger")

library(tidyverse)
library(tidymodels)

# ---- Simulate the same dataset as Exercise 3 ----
set.seed(2025)
n_samples <- 200; n_genes <- 50; n_de_genes <- 10
sample_class <- factor(rep(c("tumor", "normal"), each = n_samples / 2))
expression_matrix <- matrix(
  rnorm(n_samples * n_genes, mean = 8, sd = 1.5),
  nrow = n_samples, ncol = n_genes
)
colnames(expression_matrix) <- paste0("gene_", 1:n_genes)
for (i in 1:n_de_genes) {
  expression_matrix[1:(n_samples / 2), i] <-
    expression_matrix[1:(n_samples / 2), i] + rnorm(n_samples / 2, mean = 2, sd = 0.5)
}
gene_data <- as.data.frame(expression_matrix) |> mutate(class = sample_class)

# ---- Train/test split (same as Exercise 3) ----
set.seed(123)
split <- initial_split(gene_data, prop = 0.75, strata = class)
train_data <- training(split)
test_data  <- testing(split)

cat("Training samples:", nrow(train_data), "\n")
```

---

# 1 -- Set Up Cross-Validation Folds

Instead of evaluating on a single test set, cross-validation splits the **training data**
into multiple folds. Each fold takes a turn as the validation set while the remaining
folds are used for training.

### YOUR TURN 4.1

Create 5-fold cross-validation folds from the training data, stratified by class.

```{r}
set.seed(42)

cv_folds <- vfold_cv(
  data   = ___,
  v      = ___,
  strata = ___
)

cv_folds
```

> **CHECKPOINT:** You should see a tibble with 5 rows, each containing a different
> train/validation split.

---

# 2 -- Define Two Models

We compare logistic regression (simple, linear) against random forest (complex,
non-linear). Switching models in tidymodels is as easy as changing the specification.

### YOUR TURN 4.2

Complete both model specifications.

```{r}
# Model 1: Logistic Regression
logreg_spec <- logistic_reg() |>
  set_mode("___") |>
  set_engine("glm")

# Model 2: Random Forest
rf_spec <- rand_forest(trees = 500) |>
  set_mode("___") |>
  set_engine("ranger")
```

---

# 3 -- Define a Shared Recipe

Both models will use the same preprocessing recipe, ensuring a fair comparison.

```{r}
shared_recipe <- recipe(class ~ ., data = train_data)
```

---

# 4 -- Create Workflows and Evaluate with CV

### YOUR TURN 4.3

Create a workflow for each model, then use `fit_resamples()` to evaluate across folds.

```{r}
# Define which metrics we want
eval_metrics <- metric_set(accuracy, sensitivity, specificity, f_meas)

# ---- Logistic Regression ----
logreg_wf <- workflow() |>
  add_recipe(shared_recipe) |>
  add_model(___)

logreg_cv_results <- logreg_wf |>
  fit_resamples(
    resamples = ___,
    metrics   = eval_metrics
  )

# ---- Random Forest ----
rf_wf <- workflow() |>
  add_recipe(shared_recipe) |>
  add_model(___)

rf_cv_results <- rf_wf |>
  fit_resamples(
    resamples = ___,
    metrics   = eval_metrics
  )
```

---

# 5 -- Compare Results

### YOUR TURN 4.4

Extract and compare the mean metrics across folds.

```{r}
logreg_metrics <- logreg_cv_results |>
  collect_metrics() |>
  mutate(model = "Logistic Regression")

rf_metrics <- rf_cv_results |>
  collect_metrics() |>
  mutate(model = "Random Forest")

comparison <- bind_rows(logreg_metrics, rf_metrics) |>
  select(model, .metric, mean, std_err)

comparison
```

### YOUR TURN 4.5

Visualize the comparison with error bars.

```{r}
ggplot(comparison, aes(x = .metric, y = mean, fill = ___)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = mean - std_err, ymax = mean + std_err),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  theme_minimal() +
  labs(title = "Model Comparison: Logistic Regression vs. Random Forest",
       x = "Metric", y = "Value", fill = "Model") +
  coord_cartesian(ylim = c(0.5, 1.0))
```

> **CHECKPOINT:** Both models should perform well (accuracy > 0.80).
> Standard errors tell you how much performance varies across folds.

---

# 6 -- Interpret the Results

**Discussion questions:**

1. Which model has higher mean accuracy? Is the difference large compared to the
   standard error bars? If the error bars overlap, can you confidently say one model
   is better?

2. Look at **sensitivity** (ability to detect tumor) and **specificity** (ability to identify
   normal). Are they balanced? What would happen if you optimized only accuracy?

3. Logistic regression is linear and random forest is non-linear. For this simulated
   dataset, do you think the relationship between expression and class is linear or
   non-linear? Does the comparison support your answer?

4. We used 5 folds. What would change with 10 folds? With leave-one-out?
   What are the trade-offs?

---

# 7 -- (Stretch Goal) Repeated Cross-Validation

Repeating CV with different fold assignments gives more robust estimates.

### YOUR TURN 4.6 (Optional)

Create repeated 5-fold CV with 3 repeats and rerun the random forest.

```{r}
set.seed(42)

repeated_folds <- vfold_cv(
  data    = train_data,
  v       = 5,
  repeats = ___,
  strata  = class
)

rf_repeated_results <- rf_wf |>
  fit_resamples(
    resamples = ___,
    metrics   = eval_metrics
  )

rf_repeated_results |> collect_metrics()
```

---

# Summary

You practiced:

- Creating cross-validation folds with `vfold_cv()` and stratification
- Defining multiple model specifications and workflows
- Evaluating models with `fit_resamples()` and custom metric sets
- Comparing models using mean metrics and standard errors

**Key takeaway:** Never compare models based on a single train/test split. Always use
cross-validation. A model 1% better in accuracy but with 5% standard error is not
meaningfully better.

---

# Hints

- 4.1: Use `train_data`, 5 folds, stratify by `class`.
- 4.2: Both models use `classification` mode.
- 4.3: Add `logreg_spec` / `rf_spec`. Use `cv_folds` as resamples.
- 4.5: Fill by `model`.
- 4.6: Use 3 repeats and `repeated_folds` as resamples.

---

*Next up: Exercise 5 -- Tuning and Final Evaluation*
