---
title: "Exercise 2: Unsupervised Exploration"
subtitle: "Machine Learning with R in the Life Sciences"
author: "Konstantin Pelz, Prof. Dr. Markus List"
date: "2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Goal

You will run **PCA** and **k-means clustering** on the Palmer Penguins data, then evaluate
how well these unsupervised methods recover the known species labels -- without
ever telling the algorithm which species each penguin belongs to.

**Time:** ~15 minutes

**Corresponds to:** Unsupervised Learning section of the lecture

---

# 0 -- Setup

```{r}
if (!require(palmerpenguins)) install.packages("palmerpenguins")
if (!require(tidyverse))      install.packages("tidyverse")

library(palmerpenguins)
library(tidyverse)

# Prepare clean data (same as Exercise 1)
penguins_clean <- penguins |>
  drop_na() |>
  select(species, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)
```

# 1 -- Principal Component Analysis (PCA)

PCA reduces high-dimensional data to a few components that capture the most variance.
Before running PCA, we need to select only the **numeric** columns and **scale** them
(so features measured in different units are comparable).

### YOUR TURN 2.1

Select only the numeric columns and run PCA with `prcomp()`. The arguments `center`
and `scale.` ensure that each feature is standardized before the analysis.

```{r}
# Select only numeric columns (remove species)
penguins_numeric <- penguins_clean |>
  select(-___)

# Run PCA
pca_result <- prcomp(penguins_numeric, center = TRUE, scale. = ___)
```

### YOUR TURN 2.2

Check how much variance each principal component explains using `summary()`.

```{r}
summary(___)
```

> **CHECKPOINT:** PC1 should explain roughly **69%** of the total variance, and PC1 + PC2
> together should explain roughly **88%**.

**Discussion:** Why is it useful to know how much variance each PC captures?

---

# 2 -- Visualize PCA Results

Now let us plot the first two PCs and color the points by their known species label.

### YOUR TURN 2.3

Complete the code to create a PCA scatter plot.

```{r}
# Extract PC scores and combine with species labels
pca_data <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  species = penguins_clean$species
)

# Create the scatter plot
ggplot(pca_data, aes(x = ___, y = ___, color = ___)) +
  geom_point(alpha = 0.7, size = 2) +
  theme_minimal() +
  labs(title = "PCA of Penguin Measurements",
       x = paste0("PC1 (", round(summary(pca_result)$importance[2, 1] * 100, 1), "% variance)"),
       y = paste0("PC2 (", round(summary(pca_result)$importance[2, 2] * 100, 1), "% variance)"))
```

**Discussion:** Do the species form distinct groups in PC space? Which species is most
separated from the others?

---

# 3 -- K-Means Clustering

K-means groups data points into *k* clusters based on their similarity (distance).
The algorithm does NOT know the true species labels.

### YOUR TURN 2.4

Run k-means with k=3 (because we know there are 3 species).

```{r}
set.seed(42)  # For reproducibility

kmeans_result <- kmeans(
  scale(penguins_numeric),  # Scale the data
  centers = ___,            # Number of clusters
  nstart = 25               # Run 25 times, keep best
)
```

### YOUR TURN 2.5

Visualize the k-means clusters on the PCA plot and compare them to the true species.

```{r}
pca_data$cluster <- as.factor(kmeans_result$cluster)

# Plot clusters
ggplot(pca_data, aes(x = PC1, y = PC2, color = ___, shape = species)) +
  geom_point(alpha = 0.7, size = 2.5) +
  theme_minimal() +
  labs(title = "K-Means Clusters vs. True Species",
       color = "Cluster", shape = "True Species")
```

---

# 4 -- Evaluate the Clustering

### YOUR TURN 2.6

Create a confusion-style table to see how well the clusters match the true species.

```{r}
table(True_Species = penguins_clean$___, Cluster = kmeans_result$___)
```

> **CHECKPOINT:** One cluster should map almost perfectly to Gentoo, while
> Adelie and Chinstrap are mostly separated but with some mixing.

---

# 5 -- What Happens with the Wrong k?

### YOUR TURN 2.7

Run k-means with k=2 instead and visualize the result.

```{r}
set.seed(42)
kmeans_k2 <- kmeans(scale(penguins_numeric), centers = ___, nstart = 25)

pca_data$cluster_k2 <- as.factor(kmeans_k2$cluster)

ggplot(pca_data, aes(x = PC1, y = PC2, color = cluster_k2, shape = species)) +
  geom_point(alpha = 0.7, size = 2.5) +
  theme_minimal() +
  labs(title = "K-Means with k=2 vs. True Species",
       color = "Cluster", shape = "True Species")
```

**Discussion questions:**

1. Which two species got merged into one cluster? Does this make sense from the PCA?

2. In a real scenario where you do NOT know the true labels, how would you decide
   the right value of k? (Hint: think about the silhouette score from the lecture.)

3. This dataset has only 4 features. In transcriptomics, you might have 20,000 genes.
   Would PCA be more or less useful in that setting? Why?

---

# Summary

You practiced:

- Running PCA with `prcomp()` and interpreting variance explained
- Visualizing data in reduced dimensions (PC1 vs PC2)
- Running k-means with `kmeans()` and varying k
- Comparing unsupervised clusters to known labels
- The importance of scaling data before distance-based methods

**Key takeaway:** Unsupervised methods can discover real biological structure in data,
but they do not guarantee that the clusters match your categories of interest.

---

# Hints

- 2.1: Remove the `species` column. Set `scale.` to `TRUE`.
- 2.3: x = `PC1`, y = `PC2`, color = `species`.
- 2.4: We want 3 clusters to match the 3 species.
- 2.5: Color by `cluster` to see the k-means result; shape by species to compare.
- 2.6: Use `species` from the data and `cluster` from the k-means result.

---

*Next up: Exercise 3 -- Your First Classifier*
