---
title: "Exercise 1: Meet Your Data -- SOLUTIONS"
subtitle: Machine Learning with R in the Life Sciences
output:
  pdf_document:
    toc: false
  html_document:
    toc: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Setup

```{r}
library(palmerpenguins)
library(tidyverse)
data("penguins")
```

### 1.1
```{r}
dim(penguins)
head(penguins)
summary(penguins)
```

### 1.2
```{r}
str(penguins)
```

**Q: Which columns are numeric? Which are categorical? Which column could be a target variable (something you would want to predict)?**

**Numeric columns:** `bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`,
`body_mass_g`, and `year` (technically numeric but functions as a categorical
grouping variable here — there are only 3 unique years).

**Categorical columns:** `species`, `island`, and `sex`.

You can verify this with `str(penguins)` — numeric columns show as `num` or `int`,
categorical ones as `fct` (factor).

**Target variable:** The most natural target is `species` — predicting which species a
penguin belongs to based on its body measurements. This is a classification task
with three classes.

Other valid targets depending on the research question:

- `sex` — predicting sex from morphology (binary classification)
- `body_mass_g` — predicting body mass from the other measurements (regression)
- `island` — predicting island of origin (classification, though this is partly
  confounded with species)

The choice of target depends on the scientific question. In ML terms, the target is
always the column you want to predict, and everything else becomes a candidate
feature.

### 1.3
```{r}
penguins_clean <- penguins |>
  drop_na()
dim(penguins_clean)
```

### 1.4
```{r}
penguins_clean |>
  count(species)
```

### 1.5
```{r}
penguins_clean |>
  group_by(species) |>
  summarize(mean_mass = mean(body_mass_g))
```

### 1.6
```{r}
ggplot(penguins_clean, aes(x = species, fill = species)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Number of Penguins per Species", x = "Species", y = "Count")
```

### 1.7
```{r}
ggplot(penguins_clean, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(alpha = 0.7, size = 2) +
  theme_minimal() +
  labs(title = "Bill Dimensions by Species", x = "Bill Length (mm)", y = "Bill Depth (mm)")
```

### 1.8
```{r}
ggplot(penguins_clean, aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Flipper Length by Species", x = "Species", y = "Flipper Length (mm)")
```

### Discussion Questions — Solutions

**Q1: If we wanted to predict the species of a penguin from its measurements, which columns would be the features (inputs) and which would be the target (output)?**

The **target** (output) is the `species` column — that is what we want to predict.
The **features** (inputs) are the four numeric measurement columns:
`bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, and `body_mass_g`.

The columns `island`, `sex`, and `year` could also serve as features, but they are
categorical and would require encoding. For a first ML exercise we focus on the
numeric measurements because they directly describe the penguin's morphology
and are biologically meaningful for distinguishing species.

**Q2: Is this a classification or a regression problem? Why?**

This is a **classification** problem because the target variable `species` is
categorical (Adelie, Chinstrap, Gentoo) — we are predicting a discrete class label,
not a continuous number. If we were instead predicting `body_mass_g` from the
other measurements, *that* would be a regression problem.

**Q3: Looking at your scatter plot from 1.7: do you think a machine learning model could separate the three species? Which species pair might be hardest to distinguish?**

Yes. The scatter plot of `bill_length_mm` vs. `bill_depth_mm` shows that the three
species form relatively distinct clusters, so an ML model should be able to separate
them reasonably well.

**Adelie and Chinstrap** are the hardest pair to distinguish — they overlap
substantially in bill depth and have adjacent ranges in bill length. Gentoo penguins
form a clearly separated group (lower bill depth, higher bill length), making them
the easiest to identify.

This illustrates why we often need multiple features: adding flipper length and body
mass would help separate Adelie from Chinstrap further.

**Q4: We have 333 samples. Is this a lot or a little for ML? What if we had 33,000 genes instead of 4 numeric features?**

For 4 numeric features, 333 samples is quite reasonable — a common rule of thumb
is at least 10–20 observations per feature, and we far exceed that.

However, if we had **33,000 genes** instead of 4 features, the situation reverses
dramatically: we would have far more features than samples (p >> n). This is the
**curse of dimensionality** — the model has too many degrees of freedom relative to
the data and will overfit to noise. In that high-dimensional setting we would need:

- Feature selection (e.g., variance filtering, selecting only DE genes)
- Dimensionality reduction (PCA)
- Regularized models (Lasso, Elastic Net)

This p >> n scenario is extremely common in transcriptomics, proteomics, and other
omics data.