---
title: "Exercise 3: Your First Classifier"
subtitle: "Machine Learning with R in the Life Sciences"
author: "Konstantin Pelz, Prof. Dr. Markus List"
date: "2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Goal

You will split data into training and test sets, train your first classifier using
**tidymodels**, make predictions, and evaluate the result with a confusion matrix.
We switch to a **simulated transcriptomics** dataset: classifying tumor vs. normal
samples from gene expression data.

**Time:** ~20 minutes

**Corresponds to:** Supervised Learning + Data Splitting sections of the lecture

---

# 0 -- Setup

```{r}
if (!require(tidymodels)) install.packages("tidymodels")
if (!require(tidyverse))  install.packages("tidyverse")

library(tidyverse)
library(tidymodels)
```

# 1 -- Generate a Simulated Transcriptomics Dataset

In practice, you would load data from GEO or a similar repository. Here, we simulate
a gene expression dataset with 200 samples and 50 genes. Some genes are
differentially expressed between tumor and normal tissue.

**You do not need to change anything here** -- just run the cell and read the comments.

```{r}
set.seed(2025)

n_samples <- 200
n_genes   <- 50
n_de_genes <- 10  # 10 genes are truly differentially expressed

# Create sample labels
sample_class <- factor(rep(c("tumor", "normal"), each = n_samples / 2))

# Simulate gene expression: most genes are random noise
expression_matrix <- matrix(
  rnorm(n_samples * n_genes, mean = 8, sd = 1.5),
  nrow = n_samples, ncol = n_genes
)
colnames(expression_matrix) <- paste0("gene_", 1:n_genes)

# Make the first n_de_genes differentially expressed
for (i in 1:n_de_genes) {
  expression_matrix[1:(n_samples / 2), i] <-
    expression_matrix[1:(n_samples / 2), i] + rnorm(n_samples / 2, mean = 2, sd = 0.5)
}

# Combine into a data frame
gene_data <- as.data.frame(expression_matrix) |>
  mutate(class = sample_class)

cat("Dataset dimensions:", dim(gene_data), "\n")
cat("Class distribution:\n")
table(gene_data$class)
```

### YOUR TURN 3.1

Explore the data. Create a boxplot comparing expression of `gene_1` (differentially
expressed) between tumor and normal. Then do the same for `gene_40` (not DE).

```{r}
# Differentially expressed gene
ggplot(gene_data, aes(x = class, y = ___, fill = class)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Gene 1 (Differentially Expressed)")
```

```{r}
# Non-differentially expressed gene
ggplot(gene_data, aes(x = class, y = ___, fill = class)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Gene 40 (Not Differentially Expressed)")
```

**Discussion:** Can you visually see the difference in gene_1? What about gene_40?

---

# 2 -- Split the Data

We need to split our data into **training** (to learn from) and **test** (to evaluate on).
The test set must remain untouched until the very end.

### YOUR TURN 3.2

Use `initial_split()` to create a 75/25 split. The `strata` argument ensures that both
sets have the same proportion of tumor and normal samples.

```{r}
set.seed(123)

split <- initial_split(
  data   = gene_data,
  prop   = ___,
  strata = ___
)

train_data <- training(split)
test_data  <- testing(split)

cat("Training samples:", nrow(train_data), "\n")
cat("Test samples:", nrow(test_data), "\n")
cat("\nTraining class distribution:\n")
table(train_data$class)
```

> **CHECKPOINT:** Training should have 150 samples, test should have 50.

---

# 3 -- Define a Model

We train a **logistic regression** -- simple, fast, and interpretable.

### YOUR TURN 3.3

Complete the model specification. We want classification mode.

```{r}
log_reg_spec <- logistic_reg() |>
  set_mode("___") |>
  set_engine("glm")
```

---

# 4 -- Create a Workflow and Train

A **workflow** bundles the preprocessing (recipe) and the model together.

### YOUR TURN 3.4

Create a recipe, combine with the model, and fit to training data.

```{r}
# Recipe: predict class from everything else
train_recipe <- recipe(___ ~ ., data = train_data)

# Workflow: combine recipe + model
wf <- workflow() |>
  add_recipe(___) |>
  add_model(___)

# Fit to training data
fitted_model <- wf |>
  fit(data = ___)

fitted_model
```

---

# 5 -- Predict on the Test Set

### YOUR TURN 3.5

Use the fitted model to make predictions on the test data.

```{r}
predictions <- fitted_model |>
  predict(new_data = ___) |>
  bind_cols(test_data |> select(class))

head(predictions)
```

---

# 6 -- Evaluate with a Confusion Matrix

### YOUR TURN 3.6

Create a confusion matrix to see how many samples were correctly classified.

```{r}
conf_mat(predictions, truth = ___, estimate = ___)
```

### YOUR TURN 3.7

Calculate accuracy and balanced accuracy.

```{r}
metrics_result <- predictions |>
  metrics(truth = class, estimate = .pred_class)

metrics_result
```

> **CHECKPOINT:** Your accuracy should be approximately **85-95%**. If it is much lower,
> check that you used the correct column names.

**Discussion questions:**

1. Is accuracy alone sufficient to judge this model? What if the dataset were 95%
   normal and 5% tumor?

2. Looking at the confusion matrix: does the model make more false positives or
   false negatives? Which error is worse in a clinical screening context?

3. We used ALL 50 genes as features. Some are just noise. Would the model
   perform better with fewer, carefully selected genes?

---

# Summary

You practiced:

- Splitting data with `initial_split()` and stratification
- Defining a model with `logistic_reg()` and `set_mode()`
- Creating workflows with `recipe()`, `workflow()`, `add_recipe()`, `add_model()`
- Fitting, predicting, and evaluating
- The tidymodels pattern: recipe + model + workflow + fit + predict + evaluate

**Key takeaway:** This workflow pattern is the same for ANY model. Only the model
specification changes.

---

# Hints

- 3.1: Use `gene_1` and `gene_40` for the y-axis.
- 3.2: Use `prop = 0.75` and `strata = class`.
- 3.3: The mode is `classification` (not regression).
- 3.4: The target is `class`. Add `train_recipe` and `log_reg_spec` to the workflow. Fit on `train_data`.
- 3.5: Predict on `test_data`.
- 3.6: truth = `class`, estimate = `.pred_class`.

---

*Next up: Exercise 4 -- Model Comparison with Cross-Validation*
