---
title: "Exercise 5: Tuning and Final Evaluation"
subtitle: "Machine Learning with R in the Life Sciences"
author: "Konstantin Pelz, Prof. Dr. Markus List"
date: "2025"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Goal

Your model from Exercise 4 used default hyperparameters. In this exercise you will
**tune** the random forest hyperparameters to find the best configuration, then
perform a **final evaluation** on the held-out test set, including an ROC curve and
feature importance.

**Time:** ~20 minutes

**Corresponds to:** Model Improvement + Result Evaluation sections of the lecture

---

# 0 -- Setup and Data

```{r}
if (!require(tidymodels)) install.packages("tidymodels")
if (!require(tidyverse))  install.packages("tidyverse")
if (!require(ranger))     install.packages("ranger")
if (!require(vip))        install.packages("vip")

library(tidyverse)
library(tidymodels)
library(vip)

# ---- Same simulated data as Exercises 3-4 ----
set.seed(2025)
n_samples <- 200; n_genes <- 50; n_de_genes <- 10
sample_class <- factor(rep(c("tumor", "normal"), each = n_samples / 2))
expression_matrix <- matrix(
  rnorm(n_samples * n_genes, mean = 8, sd = 1.5),
  nrow = n_samples, ncol = n_genes
)
colnames(expression_matrix) <- paste0("gene_", 1:n_genes)
for (i in 1:n_de_genes) {
  expression_matrix[1:(n_samples / 2), i] <-
    expression_matrix[1:(n_samples / 2), i] + rnorm(n_samples / 2, mean = 2, sd = 0.5)
}
gene_data <- as.data.frame(expression_matrix) |> mutate(class = sample_class)

set.seed(123)
split <- initial_split(gene_data, prop = 0.75, strata = class)
train_data <- training(split)
test_data  <- testing(split)

# CV folds for tuning
set.seed(42)
cv_folds <- vfold_cv(train_data, v = 5, strata = class)
```

---

# 1 -- Define a Tunable Model

In tidymodels, you mark hyperparameters for tuning using `tune()` as a placeholder.

### YOUR TURN 5.1

Create a random forest with tunable hyperparameters:

- `mtry`: Number of features randomly sampled at each split
- `trees`: Number of trees in the forest
- `min_n`: Minimum number of samples in a leaf node

```{r}
rf_tunable <- rand_forest(
  mtry  = ___,
  trees = ___,
  min_n = ___
) |>
  set_mode("classification") |>
  set_engine("ranger", importance = "impurity")
```

---

# 2 -- Create a Tuning Grid

A grid defines which hyperparameter combinations to try.

### YOUR TURN 5.2

Create a regular grid. We limit mtry to between 3 and 20, trees to a few options,
and min_n to small values.

```{r}
rf_grid <- grid_regular(
  mtry(range = c(3, 20)),
  trees(range = c(200, 800)),
  min_n(range = c(2, 10)),
  levels = ___
)

cat("Number of combinations to try:", nrow(rf_grid), "\n")
rf_grid
```

---

# 3 -- Run the Tuning

### YOUR TURN 5.3

Create a workflow and run `tune_grid()` to evaluate all combinations across CV folds.

```{r}
rf_wf <- workflow() |>
  add_recipe(recipe(class ~ ., data = train_data)) |>
  add_model(___)

tune_results <- rf_wf |>
  tune_grid(
    resamples = ___,
    grid      = ___,
    metrics   = metric_set(accuracy, f_meas, roc_auc)
  )
```

> **Note:** This may take 30-60 seconds depending on your computer.

---

# 4 -- Inspect the Tuning Results

### YOUR TURN 5.4

View the top 5 hyperparameter combinations ranked by F1 score.

```{r}
tune_results |>
  show_best(metric = "___", n = 5)
```

### YOUR TURN 5.5

Visualize how performance changes with different hyperparameter values.

```{r}
autoplot(tune_results, metric = "f_meas")
```

**Discussion:** Which hyperparameter seems to matter most? Does performance
plateau after a certain number of trees?

---

# 5 -- Select the Best Model and Finalize

### YOUR TURN 5.6

Select the best hyperparameters and finalize the workflow.

```{r}
best_params <- tune_results |>
  select_best(metric = "___")

cat("Best hyperparameters:\n")
best_params

# Finalize the workflow with the best parameters
final_wf <- rf_wf |>
  finalize_workflow(___)
```

---

# 6 -- Final Evaluation on the Test Set

Now we train the finalized model on ALL training data and evaluate on the held-out
test set that the model has never seen.

### YOUR TURN 5.7

Use `last_fit()` to do the final train + evaluate in one step.

```{r}
final_fit <- final_wf |>
  last_fit(split = ___)

# View test set metrics
final_fit |> collect_metrics()
```

> **CHECKPOINT:** Test set accuracy should be in the range of 0.80 to 0.95.

---

# 7 -- ROC Curve

### YOUR TURN 5.8

Plot the ROC curve on the test set.

```{r}
test_predictions <- final_fit |> collect_predictions()

test_predictions |>
  roc_curve(truth = ___, .pred_normal) |>
  autoplot() +
  labs(title = "ROC Curve on Test Set")
```

### YOUR TURN 5.9

Create a confusion matrix for the test set.

```{r}
test_predictions |>
  conf_mat(truth = ___, estimate = ___)
```

---

# 8 -- Feature Importance

Random forests tell you which features (genes) contributed most to predictions.

### YOUR TURN 5.10

Extract the fitted model and plot feature importance for the top 15 genes.

```{r}
final_model <- final_fit |>
  extract_fit_parsnip()

final_model |>
  vip(num_features = ___) +
  theme_minimal() +
  labs(title = "Top Features (Genes) by Importance")
```

> **CHECKPOINT:** The top features should mostly be gene_1 through gene_10 (the truly
> differentially expressed ones). If noise genes appear in the top 10, the model
> might be picking up on random patterns.

**Discussion questions:**

1. Do the most important genes match the truly DE genes (gene_1 to gene_10)?
   What does it mean if a noise gene appears as important?

2. We used `importance = "impurity"` (Gini importance). The handout describes
   permutation importance. What is the difference?

3. In a real experiment, you would not know which genes are truly informative.
   How could you use feature importance to generate biological hypotheses?

4. We tuned 27 combinations. Would 1000 always give a better model?
   What are the risks?

---

# Summary

You practiced:

- Marking hyperparameters for tuning with `tune()`
- Creating tuning grids with `grid_regular()`
- Running hyperparameter search with `tune_grid()`
- Selecting the best model with `select_best()` and `finalize_workflow()`
- Final evaluation with `last_fit()`
- Plotting ROC curves and confusion matrices
- Extracting and interpreting feature importance with `vip()`

**Key takeaway:** The full ML pipeline is: split data, define model, set up CV, tune
hyperparameters, select best, evaluate ONCE on test set, interpret results.
The test set is only used at the very end.

---

# Hints

- 5.1: Replace each `___` with `tune()`.
- 5.2: Use `levels = 3`. This creates 3 x 3 x 3 = 27 combinations.
- 5.3: Add `rf_tunable`. Use `cv_folds` and `rf_grid`.
- 5.4: Use `f_meas` (this is F1 score in tidymodels).
- 5.6: Metric is `f_meas`. Pass `best_params` to `finalize_workflow()`.
- 5.7: Pass the original `split` object (not `cv_folds`).
- 5.8: truth = `class`.
- 5.9: truth = `class`, estimate = `.pred_class`.
- 5.10: Use `15` for num_features.

---

*Next up: Bonus Exercise -- Spot the Leakage*
