---
title: "Exercise 4: Model Comparison -- SOLUTIONS"
subtitle: Machine Learning with R in the Life Sciences
output:
  pdf_document:
    toc: false
  html_document:
    toc: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Setup

```{r}
library(tidyverse)
library(tidymodels)

set.seed(2025)
n_samples <- 200; n_genes <- 50; n_de_genes <- 10
sample_class <- factor(rep(c("tumor", "normal"), each = n_samples / 2))
expression_matrix <- matrix(rnorm(n_samples * n_genes, mean = 8, sd = 1.5),
                            nrow = n_samples, ncol = n_genes)
colnames(expression_matrix) <- paste0("gene_", 1:n_genes)
for (i in 1:n_de_genes) {
  expression_matrix[1:(n_samples / 2), i] <-
    expression_matrix[1:(n_samples / 2), i] + rnorm(n_samples / 2, mean = 2, sd = 0.5)
}
gene_data <- as.data.frame(expression_matrix) |> mutate(class = sample_class)

set.seed(123)
split <- initial_split(gene_data, prop = 0.75, strata = class)
train_data <- training(split)
test_data  <- testing(split)
```

### 4.1
```{r}
set.seed(42)
cv_folds <- vfold_cv(data = train_data, v = 5, strata = class)
cv_folds
```

### 4.2
```{r}
logreg_spec <- logistic_reg() |> set_mode("classification") |> set_engine("glm")
rf_spec <- rand_forest(trees = 500) |> set_mode("classification") |> set_engine("ranger")
```

### 4.3
```{r}
shared_recipe <- recipe(class ~ ., data = train_data)
eval_metrics <- metric_set(accuracy, sensitivity, specificity, f_meas)

logreg_wf <- workflow() |> add_recipe(shared_recipe) |> add_model(logreg_spec)
logreg_cv_results <- logreg_wf |> fit_resamples(resamples = cv_folds, metrics = eval_metrics)

rf_wf <- workflow() |> add_recipe(shared_recipe) |> add_model(rf_spec)
rf_cv_results <- rf_wf |> fit_resamples(resamples = cv_folds, metrics = eval_metrics)
```

### 4.4
```{r}
logreg_metrics <- logreg_cv_results |> collect_metrics() |> mutate(model = "Logistic Regression")
rf_metrics <- rf_cv_results |> collect_metrics() |> mutate(model = "Random Forest")
comparison <- bind_rows(logreg_metrics, rf_metrics) |> select(model, .metric, mean, std_err)
comparison
```

### 4.5
```{r}
ggplot(comparison, aes(x = .metric, y = mean, fill = model)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err),
                position = position_dodge(width = 0.8), width = 0.2) +
  theme_minimal() +
  labs(title = "Model Comparison", x = "Metric", y = "Value", fill = "Model") +
  coord_cartesian(ylim = c(0.5, 1.0))
```

**Q1: Which model has higher mean accuracy? Is the difference large compared to the standard error bars? If the error bars overlap, can you confidently say one model is better?**

*(The exact numbers vary by run, but typically random forest and logistic regression
perform similarly on this dataset, with random forest having a slight edge.)*

If the error bars **overlap**, you **cannot confidently say** one model is better. The
overlap means the difference in mean accuracy is within the range of normal
variation across CV folds. In statistical terms, the difference is not significant.

This is actually a common and important result: for many datasets, especially
smaller ones, several different models perform comparably. In that situation, you
should prefer the **simpler, more interpretable model** (here: logistic regression)
because:

- It is easier to explain and debug.
- It is less likely to overfit.
- The coefficients have a direct biological interpretation (log-odds).

Only choose a more complex model when it shows a clear, consistent advantage
beyond the noise of cross-validation.

**Q2: Look at sensitivity (ability to detect tumor) and specificity (ability to identify normal). Are they balanced? What would happen if you optimized only accuracy?**

With our balanced simulated dataset (50/50 tumor/normal), sensitivity and
specificity should be roughly similar for both models. This is the easy case.

If you **optimized only accuracy** on an **imbalanced** dataset (say 90% normal,
10% tumor), the model could learn to predict "normal" almost always and still
achieve 90% accuracy. The result: high specificity (correctly identifies most normals)
but very low sensitivity (misses most tumors). This is clinically dangerous because
the whole point of a screening model is to catch the rare disease cases.

That is why in practice you should:

- Always look at sensitivity and specificity separately, not just accuracy.
- Use metrics like balanced accuracy, F1, or AUC that penalize this shortcut.
- Consider the application context: in medical screening, sensitivity is usually
  prioritized over specificity.

**Q3: Logistic regression is linear and random forest is non-linear. For this simulated dataset, do you think the relationship between expression and class is linear or non-linear? Does the comparison support your answer?**

In our simulation, the DE genes have a **shifted mean** between tumor and normal
(we added a constant to gene_1 through gene_10 in the tumor group). This creates
a **linear** relationship — the classes can be separated by a straight boundary in
feature space. There are no interaction effects, threshold effects, or complex
nonlinear patterns.

The comparison supports this: logistic regression performs **nearly as well as** (or
equal to) random forest. If the true relationship were strongly nonlinear (e.g.,
tumor samples had high expression of gene_1 *only when* gene_2 was also low),
random forest would show a clear advantage because it can model interactions and
nonlinear boundaries, while logistic regression cannot.

**Takeaway:** When the signal is linear, a linear model is sufficient — and preferable
for its interpretability. Random forests and other nonlinear methods earn their keep
when relationships in the data are complex, which is often the case with real
biological data (pathway interactions, epistasis, threshold effects).

**Q4: We used 5 folds. What would change with 10 folds? With leave-one-out? What are the trade-offs?**

| | 5-fold CV | 10-fold CV | Leave-one-out (LOOCV) |
|---|---|---|---|
| **Training set size** | 80% of data | 90% of data | n - 1 samples |
| **Bias** | Slightly higher (smaller training set) | Lower | Lowest (almost all data used for training) |
| **Variance** | Lower (fewer, more different folds) | Moderate | Highest (folds are nearly identical, estimates are correlated) |
| **Computation** | Fast (5 fits) | Moderate (10 fits) | Expensive (n fits, one per sample) |
| **Stability** | More variable between runs | More stable | Deterministic (no randomness) |

**Practical guidance:**

- **5-fold CV** is a good default for smaller datasets — fast and provides a reasonable
  estimate. It slightly underestimates performance because each fold trains on only
  80% of the data.
- **10-fold CV** is a common standard in ML literature and generally offers a better
  bias-variance tradeoff. It is often paired with **repeats** (e.g., 5 × repeated 10-fold
  CV) to further stabilize estimates.
- **LOOCV** uses the most data for training (lowest bias) but produces highly
  correlated fold estimates, leading to high variance in the final metric. It is also
  computationally expensive. It is mainly useful for very small datasets (n < 50)
  where you cannot afford to hold out even 10% of the data.

For this course's dataset (200 samples), 5-fold or 10-fold with repeats are both
sensible choices.

### 4.6 (Stretch)
```{r}
set.seed(42)
repeated_folds <- vfold_cv(data = train_data, v = 5, repeats = 3, strata = class)
rf_repeated_results <- rf_wf |> fit_resamples(resamples = repeated_folds, metrics = eval_metrics)
rf_repeated_results |> collect_metrics()
```
