---
title: "Exercise 5: Tuning and Final Evaluation -- SOLUTIONS"
subtitle: Machine Learning with R in the Life Sciences
output:
  pdf_document:
    toc: false
  html_document:
    toc: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Setup

```{r}
library(tidyverse)
library(tidymodels)
library(vip)

set.seed(2025)
n_samples <- 200; n_genes <- 50; n_de_genes <- 10
sample_class <- factor(rep(c("tumor", "normal"), each = n_samples / 2))
expression_matrix <- matrix(rnorm(n_samples * n_genes, mean = 8, sd = 1.5),
                            nrow = n_samples, ncol = n_genes)
colnames(expression_matrix) <- paste0("gene_", 1:n_genes)
for (i in 1:n_de_genes) {
  expression_matrix[1:(n_samples / 2), i] <-
    expression_matrix[1:(n_samples / 2), i] + rnorm(n_samples / 2, mean = 2, sd = 0.5)
}
gene_data <- as.data.frame(expression_matrix) |> mutate(class = sample_class)

set.seed(123)
split <- initial_split(gene_data, prop = 0.75, strata = class)
train_data <- training(split)
test_data  <- testing(split)

set.seed(42)
cv_folds <- vfold_cv(train_data, v = 5, strata = class)
```

### 5.1
```{r}
rf_tunable <- rand_forest(mtry = tune(), trees = tune(), min_n = tune()) |>
  set_mode("classification") |>
  set_engine("ranger", importance = "impurity")
```

### 5.2
```{r}
rf_grid <- grid_regular(
  mtry(range = c(3, 20)),
  trees(range = c(200, 800)),
  min_n(range = c(2, 10)),
  levels = 3
)
cat("Combinations:", nrow(rf_grid), "\n")
```

### 5.3
```{r}
rf_wf <- workflow() |>
  add_recipe(recipe(class ~ ., data = train_data)) |>
  add_model(rf_tunable)

tune_results <- rf_wf |>
  tune_grid(resamples = cv_folds, grid = rf_grid,
            metrics = metric_set(accuracy, f_meas, roc_auc))
```

### 5.4
```{r}
tune_results |> show_best(metric = "f_meas", n = 5)
```

### 5.5
```{r}
autoplot(tune_results, metric = "f_meas")
```

**Q: Which hyperparameter seems to matter most? Does performance plateau after a certain number of trees?**

From the tuning results (use `autoplot(tune_results)` to visualize), students should
observe:

**`mtry` (number of features considered at each split)** is typically the most
influential hyperparameter. Setting it too low means trees cannot access the best
features at each split; setting it too high reduces the diversity between trees
(defeating the purpose of the "random" in random forest). For classification, the
default of square-root(p) (about 7 for our 50 genes) is usually a good starting point.

**`min_n` (minimum node size)** has a moderate effect. Very small values (1–2) allow
trees to grow deep and potentially overfit; very large values (20+) create shallow
trees that underfit. The effect is often less dramatic than `mtry`.

**`trees` (number of trees)** typically shows a clear **plateau**. Performance improves
as you go from a small number (50–100) to a moderate number (300–500), but
adding more trees beyond that yields diminishing returns. Going from 500 to 2000
trees rarely changes accuracy meaningfully — it just makes training slower.

This plateau happens because the ensemble's averaging effect stabilizes: once you
have enough diverse trees, adding more does not provide new information. Unlike
`mtry` and `min_n`, you generally cannot overfit by increasing the number of trees
(more trees just converges to a stable estimate), so the only cost is computation
time.

**Practical takeaway:** Focus your tuning budget on `mtry` and `min_n`. For `trees`,
pick a value where performance has plateaued (typically 500–1000) and leave it
fixed.

### 5.6
```{r}
best_params <- tune_results |> select_best(metric = "f_meas")
best_params
final_wf <- rf_wf |> finalize_workflow(best_params)
```

### 5.7
```{r}
final_fit <- final_wf |> last_fit(split = split)
final_fit |> collect_metrics()
```

### 5.8
```{r}
test_predictions <- final_fit |> collect_predictions()
test_predictions |>
  roc_curve(truth = class, .pred_normal) |>
  autoplot() + labs(title = "ROC Curve on Test Set")
```

### 5.9
```{r}
test_predictions |> conf_mat(truth = class, estimate = .pred_class)
```

### 5.10
```{r}
final_model <- final_fit |> extract_fit_parsnip()
final_model |> vip(num_features = 15) +
  theme_minimal() + labs(title = "Top Features (Genes) by Importance")
```

### Discussion Questions — Solutions

**Q1: Do the most important genes match the truly DE genes (gene_1 to gene_10)? What does it mean if a noise gene appears as important?**

Ideally the top features from the importance plot should be enriched for
**gene_1 through gene_10**. In practice students will likely see that most of the
top 10–15 features are indeed from the truly DE set, but one or two noise genes
(e.g., gene_35 or gene_42) may also appear.

This happens because of **random correlations** — by chance, a noise gene may
correlate with the outcome in the training data. This is not a bug; it is a feature of
the method and an important lesson:

- Feature importance rankings are **estimates**, not ground truth.
- They should be interpreted as **hypothesis-generating**, not definitive.
- In a real experiment you would validate top features using independent data,
  literature evidence, or functional experiments.
- With small datasets, the importance of noise genes increases — another reason
  why larger sample sizes and repeated cross-validation improve reliability.

**Q2: We used `importance = "impurity"` (Gini importance). The handout describes permutation importance. What is the difference?**

| | Gini importance (`"impurity"`) | Permutation importance (`"permutation"`) |
|---|---|---|
| **How it works** | Total reduction in Gini impurity across all splits in all trees where the feature is used | After the forest is built, randomly shuffle one feature at a time and measure the drop in accuracy |
| **Speed** | Fast — computed as a byproduct of tree building | Slower — requires additional model evaluations |
| **Bias** | Biased toward features with many unique values (high cardinality). A continuous gene with many split points tends to look more important than a binary feature, even if the binary feature is more predictive. | Unbiased and generally more reliable |
| **Recommendation** | Fine for quick exploration | Preferred for publication-quality results |

Both are covered in the course slides (Random Forest — Important Concepts, slide 63).

**Q3: In a real experiment, you would not know which genes are truly informative. How could you use feature importance to generate biological hypotheses?**

Feature importance serves as a **hypothesis generation tool**, not as proof. A
practical workflow:

1. Train your model and extract the top N most important genes.
2. Look up these genes in biological databases (e.g., GeneCards, KEGG, Gene
   Ontology) to check for involvement in relevant pathways.
3. Search the literature for known associations with your phenotype.
4. Genes that are both statistically important *and* biologically plausible become
   candidates for **experimental validation** (e.g., qPCR, knockdown, independent
   cohort replication).

This is where ML connects to biology: the model tells you *which features matter*,
and your domain knowledge tells you *whether that makes biological sense*. A
surprising gene in the top list could be a novel finding — or a false positive. That is
why independent validation is essential.

**Q4: We tuned 27 combinations. Would 1000 always give a better model? What are the risks?**

Not necessarily. More combinations are **not always better** for several reasons:

1. **Computational cost:** 1000 combinations × 10 CV folds × 5 repeats = 50,000
   model fits. With large datasets or complex models this can take hours or days.

2. **Overfitting to the validation set:** The more combinations you try, the higher the
   chance that the "best" one looks good by chance on your particular CV splits, not
   because it generalizes well. This is a subtle form of overfitting at the tuning level.
   Nested cross-validation (an outer CV around the entire tuning process) protects
   against this.

3. **Diminishing returns:** Model performance often plateaus after a modest search.
   The jump from 27 to 100 combinations may help; the jump from 100 to 1000
   rarely makes a meaningful difference for random forests. The hyperparameter
   space is usually smooth enough that a coarse grid finds a near-optimal region.

4. **Random search as an alternative:** Instead of a regular grid with 1000 points, use
   `grid_random(size = 50)` — research has shown that random search finds good
   hyperparameters with far fewer evaluations than grid search, because it explores
   the space more efficiently and avoids wasting time on unimportant dimensions.

**Practical recommendation:** Start with a coarse grid (like our 27), examine the
results, and only refine in the promising region if performance seems sensitive to
the hyperparameters.
